The acetylase activity of Cdu1 protects Chlamydia effectors from degradation to regulate bacterial exit from infected cells

Version 1

Abstract

Abstract

Many cellular processes are regulated by ubiquitin-mediated proteasomal degradation. Bacterial pathogens can regulate eukaryotic proteolysis through the delivery of proteins with de-ubiquitinating (DUB) activities. The obligate intracellular pathogen Chlamydia trachomatis secretes Cdu1 (ChlaDUB1), a dual deubiquitinase and Lys-acetyltransferase, that promotes Golgi remodeling and survival of infected host cells presumably by regulating the ubiquitination of host and bacterial proteins. Here we determined that Cdu1’s acetylase but not its DUB activity is important to protect Cdu1 from ubiquitin-mediated degradation. We further identified three C. trachomatis proteins on the pathogen-containing vacuole (InaC, IpaM, and CTL0480) that required Cdu1‘s acetylase activity for protection from degradation and determined that Cdu1 and these Cdu1-protected proteins are required for optimal egress of Chlamydia from host cells. These findings highlight a non-canonical mechanism of pathogen-mediated protection of virulence factors from degradation after their delivery into host cells and the coordinated regulation of secreted effector proteins.

Introduction

Ubiquitination is a conserved and ubiquitous post-translational modification (PTM) of proteins involving the conjugation of the carboxy-terminal glycine residue of ubiquitin (Ub) to lysine residues of target proteins. Poly-ubiquitination of substrates involves further conjugation of a Ub internal lysine residue or amino-terminal methionine (M1) with a second Ub molecule. Seven internal lysines in Ub (K6, K11, K27, K29, K33, K48, K63) and M1 are utilized by Ub conjugating enzymes to form homogeneous, branched, or mixed poly-ubiquitin (polyUb) chains (Komander and Rape., 2012). PolyUb chains with different linkage types exhibit distinct structures and functions. For example, K48- and K11-linked polyUb chains exhibit a compact conformation and are substrates for 26S proteasome-mediated degradation (Varadan et al., 2004; Tenno et al., 2004; Eddins et al., 2007; Bremm et al., 2010; Saeki., 2017). In contrast, K63-linked polyUb conjugates adopt more open conformations that enable the recruitment of multiprotein complexes that regulate the function of the target protein by proteolytic independent events (Komander et al., 2009b; Weeks et al., 2009; Datta et al., 2009; Komander and Rape., 2012). Mixed and branched polyUb chains are also emerging as important regulators of physiological functions (Swatek and Komander., 2016; Ohtake and Tsuchiya., 2017).

Protein ubiquitination regulates numerous eukaryotic cell processes including protein degradation, signal transduction, cell cycle regulation, selective autophagy, the DNA damage response, and programmed cell death. Ub also plays key roles in modulating host innate immune responses to bacterial infection (Li et al., 2016), bacterial proteins, pathogen-containing vacuoles, and bacteria themselves by targeting them for Ub-mediated degradation by proteasomal or autophagic machineries (Li et al., 2016). Because the Ub system is critical for pathogen containment, many pathogens have evolved mechanisms to counteract the impact of this PTM (Vozandychova et al., 2021). For instance, bacterial deubiquitinases (DUBs) can remove Ub from ubiquitinated substrates thereby dampening inflammatory and cell-autonomous defense mechanisms (Kubori et al., 2019). Many DUBs are cysteine proteases with a catalytic Cys, a nearby His and an Asn/Asp (Komander et al., 2009a). DUBs are typically dedicated to the removal of Ub moieties and are unable to hydrolyze other Ub-like (Ubl) modifications such as SUMO or NEDD8. However, the CE clan of Ubl proteases (ULPs) can catalyze the removal of both SUMO and NEDD8 (Ronau et al., 2016). Bacterial pathogens also encode CE clan enzymes that function as DUBs, ULPs or both. For instance, Salmonella Typhimurium SseL, Escherichia coli ElaD, and Shigella flexneri ShiCE function as Ub specific proteases (Rytkönen et al., 2007; Catic et al., 2007; Pruneda et al., 2016) while RickCE from Rickettsia belli functions as a protease directed towards both Ub and NEDD8 as does SidE from Legionella pneumophila which displays mixed activities towards Ub, NEDD8, and ISG15 (Sheedlo et al., 2015; Pruneda et al., 2016). Similarly, XopD from Xanthamonas campestris and LotB from L. pneumophila are isopeptidases exhibiting cross reactivity towards both Ub and SUMO (Pruneda et al., 2016; Schubert et al., 2020). Some CE clan bacterial effectors display acetyltransferase activity. L. pneumophila LegCE, S. Typhimurium AvrA, and YopJ from Yersinia pestis function exclusively as acetyltransferases (Mittal et al., 2006; Mukherjee et al., 2006; Jones et al., 2008; Pruneda et al., 2016). In contrast, the Chlamydia trachomatis (Ct) effector Cdu1/ChlaDUB1 is a CE clan protein that exhibits both acetyltransferase and deubiquitinating activities (Misaghi et al., 2006; Pruneda et al., 2016; Fischer et al., 2017; Pruneda et al., 2018).

Ct is an obligate intracellular bacterial pathogen responsible for human diseases of significant clinical and public health importance (Haggerty et al., 2010). Ct has a biphasic developmental cycle in which the Ct infectious propagule or elementary body (EB) invades the target host cell. Upon internalization the EB transitions to the reticulate body (RB). RBs replicate by binary fission within a pathogenic vacuole (“inclusion”) and asynchronously differentiate back to EBs. In cell culture, starting at around 48 hours-post infection (hpi) Ct will egress after lysis of the host cell or by a process termed extrusion, wherein the entire inclusion is exocytosed from the infected cell (Moulder., 1991; Abdelrahman and Belland., 2005; Hybiske and Stephens., 2007; Lee et al., 2018). The effector Cdu1 was originally identified as a deneddylating and deubiquitinating enzyme and subsequently shown to exhibit in-vitro isopeptidase activity towards both Lys48 and Lys63 linked di-Ub substrates (Misaghi et al., 2006; Claessen et al., 2013; Pruneda et al., 2016; Fischer et al., 2017). Cdu1 is unique among CE clan enzymes in that it also functions as a bona fide lysine acetylase with both is acetylase and DUB activities catalyzed by the same catalytic active site (Pruneda et al., 2018). Intriguingly, Cdu1 autoacetylation is directed towards lysines unlike other CE clan acetylases that predominantly target serine and threonine residues (Pruneda et al., 2018). In infected cells, Cdu1 localizes to the inclusion membrane where it functions to stabilize the anti-apoptotic protein Mcl-1 and to promote the repositioning of Golgi ministacks around the Ct inclusion (Fischer et al., 2017; Wang et al., 2018; Pruneda et al., 2018; Kunz et al., 2019; Auer et al., 2020). However, the mechanism by which Cdu1 promotes redeployment of Golgi ministacks and any additional roles that Cdu1 may play during Ct infection of epithelial cells remains unknown.

In this study, we show that Cdu1 protects itself and three secreted Ct effectors, InaC, IpaM, and CTL0480 from targeted ubiquitination and proteasomal degradation. InaC, IpaM, and CTL0480 are members of a larger family of bacterial proteins embedded within the inclusion membrane (Inc proteins) (Bannantine et al., 2000; Rockey et al., 2002; Chen et al., 2006; Li et al., 2008; Alzhanov et al., 2009; Dehoux et al., 2011; Lutter et al., 2012; Lutter et al., 2013; Kokes et al., 2015; Weber et al., 2015). We show that Cdu1-mediated protection from degradation is independent from its DUB activity but relies upon its Lys acetylase activity. We show that Cdu1 protects InaC to promote repositioning of Golgi ministacks and formation of actin scaffolds around the Ct inclusion, and CTL0480 to promote recruitment of myosin phosphatase target subunit 1 (MYPT1) to the inclusion. In addition, we determined that Cdu1 and Cdu1-protected Incs are required for optimal extrusion of inclusions from host cells at the late stages of infection.

Results

The C. trachomatis inclusion membrane proteins InaC, IpaM, and CTL0480 are differentially ubiquitinated in the absence of Cdu1

Cdu1 is required for Golgi repositioning around the Ct inclusion (Pruneda et al., 2018; Auer et al., 2020). To understand how Cdu1 promotes Golgi redistribution, we first generated a cdu1 null strain in a Ct L2 (LGV L2 434 Bu) background by TargeTron mediated insertional mutagenesis (pDFTT3-aadA) (Lowden et al., 2015) (S. Figure 1). Loss of Cdu1 expression in the resulting L2 cdu1::GII aadA (cdu1::GII) strain was verified by western blot analysis and by indirect immunofluorescence with antibodies raised against Cdu1 (S. Figures 2A and 2B). Because cdu2 resides directly downstream of the cdu1 locus and encodes a second Ct DUB (Cdu2/ChlaDUB2) (Misaghi et al., 2006), we first determined whether the disruption of cdu1 impacted the expression of cdu2. We detected cdu1 and cdu2 transcripts in HeLa cells infected with Ct L2 but not for the juncture between cdu1 and cdu2 (S. Figure 2C). In cells infected with Ct cdu1::GII we only detected cdu2 transcripts (S. Figure 2C) confirming that cdu1 and cdu2 are not co-expressed as part of an operon in accordance with previous observations (Albrecht et al., 2010).

We hypothesized that Cdu1s’ DUB activity promoted Golgi redistribution around inclusions and that we could identify potential targets by comparing the protein ubiquitination profile of cells infected with WT or cdu1::GII strains by quantitative mass spectrometry (MS). HeLa cell were mock infected or infected with either WT L2 or cdu1::GII strains. At 24 hpi, poly-ubiquitinated proteins were enriched from lysed cells using Tandem Ubiquitin Binding Entities (TUBEs) (LifeSensors). TUBEs consist of concatenated Ub binding associated domains (UBAs) that bind to polyUb-modified proteins with nanomolar affinities. Poly-ubiquitinated proteins of both human and Ct origin were enriched and identified by quantitative LC-MS/MS analysis.

Over 2,000 non-ubiquitinated proteins co-precipitated with TUBE 1 bound proteins across all three conditions (mock, L2, and cdu1::GII infected HeLa cells) and 3 biological replicates (S. Table 1). Among these, 47 human proteins were significantly enriched in mock infected HeLa cells and 50 human proteins were significantly enriched during Ct infection (L2 and cdu1::GII) (S. Table 3, S. Figure 3) Pathway enrichment analysis revealed that proteins involved in RNA metabolism were overrepresented among co-precipitating proteins from mock infected cells (S. Table 4, S. Figure 4) while no biological pathways or processes were overrepresented in proteins enriched from infected cells (S. Figure 4). We also identified 8 TUBE1 co-precipitating Ct proteins in HeLa cells infected with L2 and cdu1::GII (S. Table 5, S. Figure 3).

TUBE 1 affinity capture lead to the identification of 43 ubiquitinated proteins (35 human proteins and 8 Ct proteins across all 3 conditions and replicates) based on the presence of peptides containing a di-glycine remnant motif (Peng et al., 2003) (S. Tables 6-8). The lack of widespread poly-ubiquitination of either human or Ct proteins in response to Ct infection (Figure 1) was surprising given that wholesale changes in protein ubiquitination has been reported during infection of HeLa cells by intracellular pathogens like S. Typhimurium (Fiskin et al., 2016). Only two human ubiquitinated proteins (ZC3H7A and DDIT4) were found to be significantly enriched in response to WT L2 infection (Figures 1A and 1C, S. Table 8) while only one human protein (MGC3121) was preferentially ubiquitinated in HeLa cells infected with the cdu1::GII mutant strain (Figures 1B and 1C, S. Table 8). In contrast three Ct proteins, InaC (K104, K107, and K149), IpaM (K29), and CTL0480 (K115) were ubiquitinated at Lys residues in the absence of Cdu1 (Figures 1B, 1C, and 1D, S. Table 8). InaC, IpaM, and CTL0480 are Ct effector proteins that localize to the inclusion membrane (Chen et al., 2006; Alzhanov et al., 2009; Lutter et al., 2013; Kokes et al., 2015). These Type 3 secretion substrates belong to a family of over 36 inclusion membrane proteins that contain a signature bi-lobal hydrophobic transmembrane domain (Bannantine et al., 2000; Rockey et al., 2002; Li et al., 2008; Dehoux et al., 2011; Lutter et al., 2012). Because Cdu1 also localizes to the inclusion membrane (Fischer et al., 2017; Wang et al., 2018; Pruneda et al., 2018; Kunz et al., 2019) we postulated that Cdu1 directly protects InaC, IpaM, and CTL0480 from ubiquitination.

The C. trachomatis inclusion membrane proteins InaC, IpaM, and CTL0480 are ubiquitinated in the absence of Cdu1.

(A-C) Volcano plots (pairwise comparisons) of the relative abundance of human and Ct ubiquitinated peptides between mock infected HeLa cells versus HeLa cells infected with L2 (24 hpi) (A), mock infected HeLa cells versus HeLa cells infected with a cdu1 null strain (cdu1::GII) (24 hpi) (B), and HeLa cells infected with L2 (24 hpi) versus HeLa cells infected with a cdu1 null strain (24 hpi) (C). All infections were performed for 24 hours. Significance values were interpolated from 3 independent biological replicates. Ubiquitinated proteins were enriched with magnetic TUBE 1 beads (binds to polyubiquitinated proteins) and peptides identified by quantitative LC MS/MS analysis. Three Ct proteins, InaC, IpaM, and CTL0480 were differentially ubiquitinated in the absence of Cdu1. (D) InaC was ubiquitinated at K104, K107, and K149, IpaM at K290, and CTL0480 at K115 in the absence of Cdu1. TMD: Transmembrane domain. Numbering corresponds to amino acids in the protein sequence of each respective inclusion membrane protein.

Cdu1 associates with InaC, IpaM, and CTL0480

We first determined if Cdu1 co-localized with InaC, IpaM, and CTL0480 at the inclusion membrane. HeLa cells were infected for 24 hours with WT L2 or L2 expressing CTL0480-Flag from its endogenous promoter and immunostained with antibodies against Cdu1, InaC, IpaM, or the Flag epitope. Both InaC and CTL0480-Flag localized throughout the inclusion membrane while IpaM was restricted to discrete microdomains as previously reported (Alzhanov et al., 2009; Dumoux et al., 2015) (Figure 2A). Cdu1 co-localized with InaC and CTL0480-Flag and with IpaM at microdomains (Figures 2A and 2B). All four antibodies specifically recognized their corresponding antigens since immunostaining for Cdu1, InaC, IpaM or the Flag epitope was not observed in Ct strains lacking Cdu1 (cdu1::GII), InaC (M407, Kokes et al., 2015) IpaM (ipaM::GII, Meier et al., 2022), or a strain that does not express CTL0480-Flag (Figure 2A).

Cdu1 associates with InaC, IpaM, and CTL0480.

(A) Co-localization of Cdu1(magenta) with endogenous InaC (green), IpaM (green), and ectopically expressed CTL0480-Flag (green) at the L2 inclusion membrane during infection of HeLa cells at 24 hpi. HeLa cells infected with an inaC null strain (M407), an ipaM null strain (ipaM::GII), and L2 pBOMB were used as antibody specificity controls. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. Images are representative of multiple images captured across three independent replicates. (B) Line scan profiles of fluorescent signal intensities displayed in (A) showing co-localization of fluorescence intensities for endogenous Cdu1 with endogenous InaC, and IpaM and with CTL0480-Flag along the L2 inclusion membrane. (C) Schematic of Cdu1-GFP(C) (L2) fusion (Cdu1-GFP) and Cdu1-GFP variants used in co-transfections of HEK 293 cells. GFP: Green fluorescent protein. TMD: Transmembrane domain. PRD: Proline rich domain. CD: Catalytic domain. FL: Full length. TMD-: Cdu1-GFP variant lacking TMD domain. CD-: Cdu1-GFP variant lacking CD domain. (D-G) Western blot analysis of GFP immunoprecipitates from HEK 293 cells co-transfected with mammalian plasmids expressing: Cdu1-GFP variants and (D) truncated 3XFlag(N)-InaC (D/UW-3/CX CT813, amino acids 96-264), (E) V5(N)-IpaM (L2, full length), (F) V5(N)-CTL0480 (L2, full length), and (G) V5(N)-CpoS (L2, full length). Vec1: Empty pOPINN-GFP vector. Vec2: Empty pDEST53 vector. Vec3: Empty pcDNATM3.1/nV5-DEST vector. Western blot images are representative from two independent experiments.

We next determined if Cdu1 can interact with InaC, IpaM, and CTL0480 by co-transfecting HEK 293 cells with vectors expressing either full length Cdu1-GFP or truncated versions of Cdu1-GFP lacking transmembrane or catalytic domains (Figure 2C), and vectors expressing Flag-InaC, V5-IpaM, and V5-CTL0480. Transfected cells were lysed and Cdu1-GFP was immunoprecipitated with antibodies against GFP. Western blot analysis of the immunoprecipitates showed that Flag-InaC, V5-IpaM, and V5-CTL0480 co-precipitated with Cdu1-GFP (Figures 2D-F). Moreover, the transmembrane domain of Cdu1 was necessary for Cdu1-GFP to interact with all three Incs (Figures 2D-F). The interaction between Cdu1-GFP and the three tagged Incs was specific since we did not detect interactions between Cdu1-GFP and a V5-tagged version of the inclusion membrane protein CpoS (Sixt et al., 2017) (Figure 2G). We conclude from these results that Cdu1 can interact with all three Incs even in the absence of infection and that interactions are facilitated by the transmembrane domain of Cdu1.

Cdu1 protects InaC, IpaM, and CTL0480 proteins from degradation during infection

We next assessed if Cdu1 was required to stabilize endogenous InaC, IpaM, and CTL0480 in infected cells. HeLa cells were infected with either WT L2 or cdu1::GII strains and at various time points in the Ct infectious cycle crude cell lysates were analyzed by western blot to assess the relative abundance of Inc proteins. At 24 hpi, we observed a decrease in InaC protein levels in cells infected with cdu1::GII relative to cells infected with WT L2 (Figure 3A). In contrast, IpaM protein levels did not decrease until 36 hpi in cdu1::GII infected cells (Figure 3B). For CTL0480, we were not able to detect CTL0480 by western blot but were successful in following CTL0480 expression by indirect immunofluorescence (Figure 3C). The relative abundance of CTL0480 at inclusion membranes, like IpaM, was not affected in cdu1::GII inclusions at 24 hpi. However, at 36 hpi, a subpopulation of cells lost CTL0480 immunoreactivity and by 48 hpi, inclusion membranes of the cdu1::GII strain were devoid of CTL0480 while CTL0480 was prominently detected at the inclusion membranes of WT L2 (Figures 3C and 3D). As controls for the specificity of antibodies used for western blots and for indirect immunofluorescence, we included cells infected with Ct lacking ipaM (ipaM::GII, Meier et al., 2022), with an inaC nonsense mutant (M407, Kokes et al., 2015), and with Ct lacking ctl0480 (ctl0480::GII, Shaw et al., 2018). Overall, our results indicate that steady state protein levels of InaC and IpaM and CTL0480 localization at the inclusion membrane, especially at late stages of infection, are dependent on Cdu1, and that Cdu1 acts at different stages in the infection cycle.

Cdu1 stabilizes InaC, IpaM, and CTL0480.

(A) Western blot analysis of InaC in HeLa cells infected for 24 hours with L2, inaC null, and cdu1 null strains. Ct Slc1 and human alpha tubulin were used to determine Ct burdens and equal loading of protein extracts respectively. Western blot images are representative of 3 independent experiments. (B) Western blot analysis of endogenous IpaM in HeLa cells infected with L2, cdu1 null, and ipaM null strains for 24, 36, and 48 hours. Western blot images are representative of 2 independent experiments. (C) Localization of CTL0480 during Ct infection of HeLa cells at 24, 36, and 48 hpi. CTL0480 signal (green) co-localizes with the inclusion membrane protein IncA (magenta) at the L2 inclusion membrane. Arrowheads highlight cdu1 null inclusions lacking CTL0480 at 36 hpi. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. (D) Quantification of CTL0480 localization at Ct inclusion membranes as shown in (C). CTL0480 fluorescent signal was used to determine the percent of inclusions displaying CTL0480 at the inclusion membranes. Representative images and quantification of CTL0480 positive inclusions are derived from Ct inclusions imaged in 10 fields across 3 independent biological replicates. p values were determined by a student paired t-test.

The acetylase activity of Cdu1 is required for Cdu1 to protect itself, InaC, and IpaM from polyubiquitination and proteasomal degradation

The crystal structures of Cdu1 bound to Ub or Coenzyme A indicated that the adenosine and phosphate groups of Coenzyme A make contact with a helix in variable region 3 (VR-3) of Cdu1 while the Ile36-patch of Ub binds to the opposite face of the same helix (Pruneda et al., 2018). Although Cdu1 catalyzes both of its DUB and acetylase (Act) activities with the same active site (Pruneda et al., 2018) the two activities of Cdu1 can be uncoupled by the amino acid substitution K268E in VR-3 which disrupts Coenzyme A binding required for Act activity and by the amino acid substitution I225A in the Ub-binding region of VR-3 required for DUB activity (Pruneda et al., 2018). These substitutions allowed us to test which of Cdu1’s enzymatic functions are required for the observed effects on protein stability. We generated Ct shuttle plasmids (pBOMB4-MCI backbone), expressing WT Cdu1, a catalytically inactive variant of Cdu1 lacking both DUB and Act activities (Cdu1C345A) (Pruneda et al., 2018), a Cdu1 DUB-deficient variant (Cdu1I225A), and a Cdu1 Act-deficient variant (Cdu1K268E). All Cdu1 constructs were expressed as 3X Flag epitope-tagged versions and from the Cdu1 endogenous promoter.

Plasmids expressing each Cdu1 variant were transformed into the cdu1::GII mutant and the resulting strains used to infect HeLa cells for 36 and 48 hours. The levels of endogenous InaC and IpaM in cell extracts of infected cells were monitored by western blot analysis. At 36 hpi, InaC protein levels drastically decreased in cells infected with cdu1 null strains transformed with empty vector or expressing the catalytic inactive variant of Cdu1 (Cdu1C345A-Flag) (Figure 4A). Likewise, IpaM protein levels diminished at 48 hpi during infection with the same strains (Figure 4B). Both InaC and IpaM protein levels were restored to wild type levels in cdu1 null strains complemented with wild type Cdu1-Flag (Figures 4A and 4B). Unexpectedly, cells infected with a cdu1 null strain ectopically expressing the DUB deficient Cdu1 variant (Cdu1I225A-Flag) displayed wild type levels of InaC and IpaM while the Act deficient variant (Cdu1K268E-Flag) did not (Figures 4A and 4B). These results suggest that the acetylase activity of Cdu1 rather than its DUB activity is required for Cdu1’s ability to stabilize InaC and IpaM proteins.

The acetylase activity of Cdu1 is required to stabilize Cdu1, InaC, and IpaM.

(A) Western blot analysis of endogenous InaC and Cdu1-Flag catalytic variants expressed from a plasmid. HeLa cells were infected for 36 hours with L2, a cdu1 null strain, and cdu1 null strains expressing wild type Cdu1-Flag and the Cdu1 variants C345A-Flag (catalytic inactive), I225A-Flag (DUB deficient), and K268E-Flag (Act deficient). Protein lysates from HeLa cells infected with M407 (inaC null) were used to control for the specificity of anti-InaC antibodies. Western blot images are representative of two independent experiments. (B) Western blot analysis of endogenous IpaM and Cdu1-FLAG variants in crude extracts of HeLa cells infected for 48 hours with the same strains as describe in (A). Infection of HeLa cells with ipaM::GII was used to test for the specificity of the anti-IpaM antibody. Western blot images are representative of two independent experiments. (C) Western blot analysis of Cdu1C345A-Flag (catalytic inactive) and Cdu1K268E-Flag (Act deficient) expressed in a cdu1 null strain or wild type L2 background after infection of HeLa cells for 24 hours. Both Cdu1 variants are stabilized by endogenous Cdu1. (D) Western blot analysis of Cdu1-Flag and Cdu1C345A-Flag expressed in a cdu1 null strain and Cdu1C345A-Flag expressed in wild type L2 following immunoprecipitation (anti-Flag) from HeLa cell extracts after infection for 24 hours and treatment with MG132 (25 μM, 5 hours). Western blot image is a representative blot from at least three independent experiments. (E) Western blot analysis of endogenous InaC and IpaM following immunoprecipitation of proteins acetylated at lysines from inclusion membrane subcellular fractions derived from HeLa cells infected with WT L2 for 24 hours. Western blot image is representative of two independent experiments. (F) Western blot analysis (Flag WB) of acetylated lysine immunoprecipitates generated from inclusion membrane subcellular fractions derived from HeLa cells infected with WT L2 strains expressing Cdu1-Flag, CTL0480-Flag, and IpaM-Flag at 40 hpi. Western blot image is representative of two independent experiments. (G) Flag WB of acetylated lysine immunoprecipitates generated from inclusion membrane subcellular fractions derived from HeLa cells infected with Rif-R L2 (M407 parental strain) and with M407 (inaC null) expressing InaC-Flag at 24 hpi. Western blot image is representative of two independent experiments. (H) The initiator methionine, Lys10, Lys126, and Lys263 of Cdu1 are acetylated during Ct infection of HeLa cells (24 hpi). One tyrosine (Y) residue and multiple serine (S) and threonine (T) residues in Cdu1 are also phosphorylated during Ct infection of HeLa cells (24 hpi). Three PX(S/T)P MAPK phosphorylation consensus sequence motifs were identified in the proline rich domain of Cdu1. Modified residues were identified by quantitative LC MS/MS analysis of immunoprecipitated Cdu1-Flag across 3 independent biological replicates. TMD: Transmembrane domain. PR: Proline rich. PSTP: PX(S/T)P motifs. Ac: Acetylation. P: Phosphorylation. Numbering corresponds to amino acids in Cdu1 protein sequence.

When we monitored the stability of each Flag-tagged Cdu1 variant, we observed that the catalytically inactive variant of Cdu1 (C345A-Flag) was destabilized (Figures 4A and 4B-Flag WB). We reasoned that Cdu1 also protects itself from being targeted for degradation in infected cells. As with InaC and IpaM, the acetylase but not the DUB activity of Cdu1 was required for Cdu1’s stability (Figures 4A and 4B). Although the C345A (DUB-, Act-) and K268E (Act-) amino acid substitutions in Cdu1 do not destabilize Cdu1 expressed in E. coli (Pruneda et al., 2018), it was possible that these substitutions impacted the expression and/or folding of Cdu1 in Chlamydia. To determine if these Cdu1 mutants were inherently unstable, we expressed Cdu1C345A-Flag and Cdu1K268E-Flag in WT L2 in the presence of endogenous Cdu1. We found that each Flag-tagged variant was stabilized (Figure 4C), indicating that endogenous Cdu1 protected the catalytically-deficient Cdu1 variants in trans. In addition, western blot analysis of immunoprecipitated Cdu1-Flag and Cdu1C345A-Flag expressed in a cdu1::GII mutant showed that while wild-type Cdu1-Flag was not modified by Lys48-linked poly-ubiquitination, Cdu1C345A-Flag was robustly modified by Lys48-linked polyUb in the presence of the proteasome inhibitor MG132 (Figure 4D). Moreover, endogenous Cdu1 protected Cdu1C345A-Flag from Lys48-linked polyUb when Cdu1C345A-Flag was expressed in a wild type L2 background (Figure 4D). These results indicate that the loss of Cdu1 activity likely leads to its Lys48-linked poly-ubiquitination and subsequent proteasome-dependent degradation.

Because Cdu1 autoacetylates itself and its Act activity is directed towards lysines (Pruneda et al., 2018) we postulated that Cdu1 may stabilize proteins from degradation by acetylating lysine residues that are potential targets of ubiquitination. We tested this hypothesis by assessing whether Cdu1, InaC, CTL0480, and IpaM are acetylated at lysines during infection. Fractions enriched for inclusion membranes were isolated by sub-cellular fractionation from HeLa cells infected with wild-type L2 (24 hpi), and proteins acetylated at lysines were immunoprecipitated. Western blot analysis of acetyl-lysine immunoprecipitates indicated that InaC and IpaM, but not the Inc protein IncA, were acetylated (Figure 4E). Western blot analysis of anti-acetyl-lysine immunoprecipitates of inclusion membrane-enriched membrane fractions of HeLa cells infected with L2 expressing Cdu1-Flag (24 hpi), CTL0480-Flag (40 hpi), and IpaM-Flag (40 hpi) also showed that all three Flag tagged effectors were acetylated at lysines (Figure 4F). We also determined that Flag-tagged InaC expressed in an M407 (inaC null) background was acetylated at lysines as determined by western blot analysis of anti-acetyl-lysine immunoprecipitates (Figure 4G). In addition, we identified acetylated forms of Cdu1 from mass spectrometric analysis of Flag immunoprecipitates derived from extracts of HeLa cells infected with L2 expressing Cdu1-Flag, (24 hpi) (Figure 4H).

We reasoned that the lysine acetylase activity of Cdu1 is a prominent mechanism by which Cdu1 protects client proteins (at 24 hpi), since loss of Cdu1 or expression of the acetylase deficient variant of Cdu1 (Cdu1K268E-Flag) leads to a marked increase in Ub immunostaining at or near the periphery of cdu1::GII inclusions (> 80% of inclusions) compared to HeLa cells infected with cdu1::GII strains complemented with wild type or DUB deficient (I225A-Flag) Cdu1 strains (Figures 5A and 5B). Given that Cdu1 appears to localize exclusively at inclusion membranes we predicted that its activity would be spatially restricted to the inclusion periphery. We tested this premise by co-infecting HeLa cells with an incA null strain (M923 (IncA R197*), Kokes et al., 2015) and the cdu1::GII strain. IncA mediates homotypic fusion of inclusion membranes and loss of IncA results in the accumulation of multiple unfused inclusions in cells infected at high MOIs (Hackstadt et al., 1999; Suchland et al., 2000; Pannekoek et al., 2005) (Figure 5C). As expected, incA mutants which retain Cdu1 activity did not accumulate Ub at or near the periphery of inclusion membranes. In HeLa cells bearing both cdu1::GII and M923 inclusions, cdu1::GII Cdu1C345A and M923 inclusions, or cdu1::GII Cdu1K268E and M923 inclusions, only the M923 inclusions were protected from ubiquitination (Figures 5C and 5D). Based on these observations we conclude that the acetylase activity of Cdu1 protects proteins in cis and that this activity is constrained to the membrane of the pathogenic vacuole consistent with previous reports (Auer et al., 2020).

The DUB activity of Cdu1 is not required for blocking the ubiquitination of inclusion membranes.

(A) Loss of Cdu1’s acetylase activity results in the accumulation of Ub (magenta signal) at the periphery of the inclusion. HeLa cells were infected for 24 hours. Antisera against the membrane protein Cap1 (green) was used to mark inclusion membranes. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. (B) Quantification of ubiquitinated inclusions as shown in (A). The Ub fluorescent signal was used to determine the number of infected cells with Ub decorated inclusions. The total number of ubiquitinated inclusions was divided by the total number of inclusions analyzed (defined by Cap1 staining). 87%, 86%, and 91% of inclusions were decorated with Ub in HeLa cells infected with a cdu1 null strain and cdu1 null strains expressing Cdu1C345A (DUB-, Act-), and Cdu1K268E (Act-) variants respectively. Representative images in (A) and quantification of ubiquitinated inclusions in (B) were obtained from inclusions imaged in 10 fields across 3 independent biological replicates for each strain. p values were determined by a student paired t-test. (C) HeLa cells co-infected with cdu1::GII (cdu1⊗) and incA null (incA⊗, M923) strains at 24 hpi. IncA-deficient inclusions do not fuse with other inclusions. In co-infected cells, Cdu1 present on the inclusion membranes of incA⊗ strains did not block ubiquitination events at or near the inclusion membranes of neighboring cdu1⊗ strains (mCherry signal from pBOMB4-MCI plasmid). DNA stained with Hoechst is shown in blue. Scale bar: 10μm. (D) Quantification of cdu1⊗ inclusions as shown in (C) in which ubiquitination events are observed in regions of cdu1⊗ inclusion membranes that are in direct apposition to incA⊗ inclusion membranes (junctions). The total number of cdu1⊗ inclusions ubiquitinated at inclusion junctions was divided by the total number of inclusions analyzed (Cap1 staining). 66% and 61% of cdu1⊗ inclusions were decorated with Ub at junctions in HeLa cells co-infected with incA⊗ and cdu1⊗ strains or with incA⊗ and a cdu1⊗ strain ectopically expressing Cdu1K268E (Act-) respectively. Representative images in (C) and quantification of cdu1⊗ inclusions ubiquitinated at junctions in (D) are derived from inclusions imaged in 6 fields across 3 independent biological replicates for each condition. p values were determined by a student paired t-test.

Cdu1 is required for F-actin assembly and Golgi ministack repositioning around the Ct inclusion, and for MYPT1 recruitment to Ct inclusions

InaC is required for Ct to assemble F-actin scaffolds and to reposition Golgi mini stacks around the peripheries of inclusion membranes (Kokes et al., 2015; Wesolowski et al., 2017; Haines et al., 2021). Because Cdu1 regulates InaC levels, we predicted that cdu1 mutants would phenocopy inaC mutants. We quantified the number of inclusions surrounded by F-actin cages at 40 hpi. In cells infected with Rif-R L2 (wild type parental strain of M407, Nguyen and Valdivia., 2012; Kokes et al., 2015), 25% of inclusions were surrounded by F-actin, consistent with previous observations (Chin et al., 2012; Kokes et al., 2015) (Figures 6A and 6E). The number of inclusions surrounded by F-actin decreased to 7% in cells infected with M407 (inaC null) and increased to 49% in HeLa cells infected with M407 complemented with wild type InaC (Figures 6A and 6E). A significant increase in inclusions surrounded by F-actin was also observed in host cells infected with cdu1::GII mutants expressing Cdu1-Flag and Cdu1I225A-Flag (DUB-) (52% and 46% respectively) (Figures 6A and 6E) while cdu1::GII mutants transformed with an empty plasmid or expressing Cdu1C345A-Flag (DUB-Act-) and Cdu1K268E-Flag (Act-) resulted in only 8%,13%, and 10% of inclusions enveloped by F-actin respectively (Figures 6A and 6E). From these observations we conclude that the acetylase activity of Cdu1 is required for Ct to promote assembly of F-actin around the Ct inclusion likely through the stabilization of InaC.

The DUB activity of Cdu1 is not required for assembly of F-actin or Golgi ministack repositioning around the Ct inclusion while Cdu1 is required for MYPT1 recruitment to the inclusion.

Representative images of F-actin (green, Alexa FluorTM Phalloidin) assembled around the Ct inclusion (magenta, anti Cdu1 and Cap1 staining) in HeLa cells infected for 40 hours. (B) Representative images of Golgi (anti GM130 staining, magenta) around Ct inclusions (arrowheads) in HeLa cells infected for 24 hours. HeLa cells were infected with an inaC null strain transformed with empty p2TK2 vector (M407 p2TK2), the wild type strain in which M407 was derived from (Rif-R L2) and M407 ectopically expressing full length InaC (M407 p2TK2-InaC). HeLa cells were also infected with L2 pBOMB, cdu1::GII pBOMB, and cdu1 null strains ectopically expressing Cdu1-Flag (Cdu1), and the Cdu1 variants C345A-Flag (DUB-, Act-), I225A-Flag (DUB-), and K268E-Flag (Act-). (C) Representative images of MYPT1 (magenta) at Ct inclusions (green, anti-IncA staining). Arrowheads represent cdu1 null inclusions with low MYPT1 signal. DNA stained with Hoechst is shown in blue in panels A-C. Scale bar: 10μm in panels A-C. (D) Quantification of Golgi dispersal around the Ct inclusion as shown in (B). The length of Golgi dispersed around each Ct inclusion imaged was measured and normalized to the perimeter length of each inclusion (% Golgi dispersion around the inclusion). Representative images in (B) and Golgi dispersal quantified in (D) were performed from inclusions imaged in 6 fields across 3 independent biological replicates. p values were determined by a student paired t-test. (E) Quantification of Ct inclusion surrounded by F-actin as shown in (A) normalized to total inclusions analyzed. Representative images in (A) and quantification of F-actin surrounding Ct inclusions in (E) were obtained from inclusions imaged in 6 fields across 3 independent biological replicates. p values were determined by one-way ANOVAs with a Student-Newman-Keuls post hoc test. (F) Quantification of MYPT1 at Ct inclusions as shown in (C). Representative images in (C) and quantification of MYPT1 recruitment in (F) were obtained from inclusions imaged in 6 fields across 3 independent replicates. p values were determined by a student paired t-test.

We next quantified Golgi dispersal in infected HeLa cells at 24 hpi, a process that is also dependent on InaC (Kokes et al., 2015; Wesolowski et al., 2017). In Hela cells infected with an inaC null strain (M407) Golgi dispersal was limited to only 26% of the Ct inclusion perimeter. In contrast, cells infected with either L2 Rif-R or with M407 complemented with wild type InaC, the Golgi is dispersed around 45% of the inclusion perimeter (Figures 6B and 6D). Similarly, Golgi dispersal around inclusions during infection with WT Ct L2 and in cdu1::GII mutants expressing wild type Cdu1-Flag or Cdu1I225A-Flag (DUB-) was 43%, 41%, and 43% respectively (Figures 6B and 6D). In HeLa cells infected with cdu1::GII and cdu1::GII strains expressing Cdu1C345A-Flag (DUB-Act-), and Cdu1K268E-Flag (Act-), Golgi repositioning was restricted to only 24%, 23%, and 23% of inclusion perimeters respectively (Figures 6B and 6D). These results confirm that both InaC and Cdu1 are required for efficient repositioning of the Golgi around the Ct inclusion as previously reported (Kokes et al., 2015; Wesolowski et al., 2017; Pruneda et al., 2018; Auer et al., 2020) and that this process is independent of Cdu1’s DUB activity but requires its acetylase activity. Moreover, our results suggest that Cdu1 promotes Golgi repositioning by protecting InaC-mediated redistribution of the Golgi around the Ct inclusion.

CTL0480 promotes recruitment of the myosin phosphatase subunit MYPT1 to the inclusion membrane where it regulates the extrusion of intact inclusions from host cells (Lutter et al., 2013; Shaw et al., 2018). Consistent with the gradual loss of CTL0480 from inclusions in cells infected with the cdu1::GII strain starting at 36 hpi (Figures 3C and 3D) we also observed a complete loss of MYPT1 recruitment to inclusions by 48 hpi (Figures 6C and 6F).

Cdu1, InaC, IpaM, and CTL0480 are required for optimal extrusion of Ct from host cells

In the absence of Cdu1, the levels of InaC, CTL0480, and IpaM decreased late in infection (36 hpi and 48 hpi, Figures 3 and 4) suggesting that a prominent role of Cdu1 is to protect these Incs from degradation late in infection. At the end of its developmental cycle, Chlamydia exits host cells by promoting cellular lysis or by extrusion of intact inclusions (Hybiske and Stephens., 2007). Ct host cell exit by extrusion is an active process requiring a remodeling of the actin cytoskeleton and the function of Inc proteins (Hybiske and Stephens., 2007; Chin et al., 2012; Lutter et al., 2013; Shaw et al., 2018; Nguyen et al., 2018). CTL0480 recruits MYPT1 (an inhibitor of Myosin II motor complexes) to the inclusion membrane which prevents premature extrusion of Ct inclusions and loss of CTL0480 leads to increased rates of extrusion by Ct from infected HeLa cells (Lutter et al., 2013; Shaw et al., 2018) (Figure 7). Actin polymerization is also required for Ct extrusion (Hybiske and Stephens., 2007; Chin et al., 2012) suggesting that InaC dependent recruitment of F-actin to the inclusion may also contribute to optimal Ct extrusion. IpaM localizes to microdomains in the inclusion membrane that are proposed to function as foci for extrusion (Nguyen et al., 2018). Based on these observations, we postulated that Cdu1-mediated protection of CTL0480, InaC, and IpaM regulates the extrusion of Ct inclusions. We quantified the number of extrusions released from infected HeLa cells at 52 hpi and observed a 60% reduction in the number of extrusions in HeLa cells infected with the cdu1::GII strain relative to cells infected with WT L2 (Figures 7A and 7B). Complementation of cdu1::GII with either wild type Cdu1-Flag or Cdu1I225A-Flag (DUB-) restored extrusion production to near wild type levels. Infection of HeLa cells with inaC (inaC::GII, Wesolowski et al., 2017), or ipaM (ipaM::GII, Meier et al., 2020) null strains led to a 75% and 58% reduction in extrusion production respectively (Figures 7A and 7B). Impaired extrusion of inclusions by cdu1::GII, inaC::GII, and ipaM::GII strains was not due to defects in inclusion biogenesis, since all three strains produced similar number of inclusions at 48 hpi relative to cells infected with WT L2 (Supp. Figure 5). We conclude that Cdu1, InaC, and IpaM function to promote optimal extrusion of Ct inclusions from host cells.

Cdu1, InaC, and IpaM are required for optimal extrusion of Ct inclusions from HeLa cells.

(A) Representative images of extrusions isolated from HeLa cell monolayers infected with Ct strains for 52 hours. Scale bar: 200 μm (B) Quantification of the number of extruded inclusions produced by infected HeLa cell monolayers. p values were determined by a student paired t-test. (C) Quantification of the size of extruded inclusions quantified in (B). Extruded inclusions varied in size among cells infected with wild type L2 (average :43 μm), ipaM mutants (average: 56 μm) and cdu1 mutants complemented wild type Cdu1 (average: 52 μm) and Cdu1I225A (DUB-) (average: 60 μm). p values were determined by one-way ANOVAs with a Student-Newman-Keuls post hoc test. Representative images in (A) and quantification of extrusion number in (B) and extrusion size in (C) are based on images obtained from 6 fields across 3 independent biological replicates.

In contrast, infection of HeLa cells with a ctl0480::GII mutant strain led to an increase in the number of extruded inclusions as previously observed (Shaw et al., 2018) (Figures 7A and 7B). Therefore, even though the Cdu1-mediated protection of InaC and IpaM is important for the extrusion of inclusions and cdu1 mutants phenocopy the loss of InaC and IpaM, the phenotypic similarities do not extend to the increased number of extruded inclusions observed in cells infected with the ctl0480::GII mutant strain (Figures 7A and 7B). We infer from these observations that functions for both InaC and IpaM in the extrusion of inclusions are epistatic to CTL0480. Extruded inclusions produced during infection of HeLa cells also varied in size with an average diameter of 40 μm (Figures 7A and 7C). Interestingly, the loss of IpaM and over expression of Cdu1-Flag and Cdu1I225A-Flag (DUB-) shifted the size distribution of extrusions toward larger extrusions (Figures 7A and 7C) suggesting that Ct regulates the size of extruded inclusions through Cdu1.

Discussion

Several Chlamydia effector proteins localize to the inclusion membrane to regulate interactions between the Chlamydia pathogenic vacuole and the host cytoskeleton, organelles, and vesicular trafficking pathways. They also modulate host cell death programs and promote Chlamydia exit from host cells (reviewed in Bugalhão and Mota., 2019). Given the central roles that Incs play in promoting Chlamydia intracellular infection, it is not surprising that they are targeted for inactivation by host cellular defenses. In response, Chlamydia has evolved mechanisms to protect Incs. In this study we show that the acetylase activity of the effector Cdu1 protects itself as well as three independent Incs; InaC, IpaM, and CTL0480, from ubiquitination and degradation (Figure 8). Interestingly, all three Incs play prominent roles in regulating the extrusion of Chlamydia inclusions from host cells (Figure 7). Observations that the encapsulation of Chlamydia within an extruded inclusion enhances survival of Chlamydia within macrophages (Zuck et al., 2017) together with the broad conservation of extrusion as an exit strategy among Chlamydia (Zuck et al., 2016) suggests that this mechanism is important for Chlamydia pathogenesis. Thus, targeting Chlamydia Incs that regulate extrusion for Ub-mediated destruction may be advantageous for the host. In this context, we hypothesize that InaC is targeted for degradation due to its role in promoting F-actin assembly around the inclusion, since F-actin is essential for extrusion (Hybiske and Stephens., 2013; Chin et al., 2012). However, other roles of InaC like its regulation of microtubule scaffolds around the inclusion cannot be ruled out (Wesolowski et al., 2017; Haines et al., 2021). CTL0480 was previously shown to function as an inhibitor of extrusions through its role in modulating the activity of myosin light chain 2 (MLC2) (Lutter et al., 2013; Shaw et al., 2018). IpaM localizes to specialized microdomains in the inclusion membrane which are also sites of enrichment for over 9 inclusion membrane proteins including CTL0480 and MrcA, both of which are required for Chlamydia extrusion (Mital et al., 2010; Lutter et al., 2013; Nguyen et al., 2018). IpaM is a third inclusion membrane protein localizing to microdomains that is required for the optimal extrusion of inclusions and lends support to a potential role for microdomains as foci for extrusion (Nguyen et al., 2018). We also find that the loss of IpaM shifted the size distribution of extrusions towards larger inclusions indicating that Ct constraints the size of extruded inclusions (Figure 7). We speculate that heterogeneity in the size of extrusions might facilitate uptake of some extrusions by innate immune cells at infected mucosal sites. Uptake of Chlamydia by host phagocytes has been proposed to promote Chlamydia LGV dissemination to distal sites in the genital tract to avoid clearance of Chlamydia by other immune cells (Zuck et al., 2017). Protection of IpaM by Cdu1 could therefore be advantageous for Chlamydia EBs within extrusions to escape to distal sites and avoid exposure to localized inflammatory responses.

A model for acetylation mediated protection of the translocated Ct effectors InaC, IpaM, and CTL0480 from degradation.

The human cellular Ub machinery targets C. trachomatis effectors, including the inclusion membrane proteins InaC, IpaM, and CTL0480 for ubiquitination and subsequent protein degradation. C. trachomatis has evolved strategies to counter such defense mechanisms by translocating the effector Cdu1. Cdu1 utilizes its acetylase activity rather than its deubiquitinase activity to protect itself and all three inclusion membrane proteins from being targeted for ubiquitination and degradation. Cdu1 protects InaC in order to promote the recruitment of F-actin scaffolds and Golgi ministacks to the inclusion perimeter and CTL0480 to recruit the Myosin II regulator MYPT1. All three inclusion proteins and Cdu1 also participate in promoting extrusion of C. trachomatis inclusions late in infection. Image created with BioRender.com.

Effectors that modulate the activity of other translocated effectors are referred to as “metaeffectors”, a term coined by Kubori and colleagues after observing that the L. pneumophila effector LubX which functions as an E3 ligase, ubiquitinates the translocated effector SidH leading to its degradation (Kubori et al., 2010). Several other effector-metaeffector interactions have been described in L. pneumophila, Salmonella enterica, and Brucella abortus which regulate the activity of other effectors either directly or indirectly by modifying the same host target or cellular process (Kubori et al., 2010; Neunuebel et al., 2011; Jeong et al., 2015; Urbanus et al., 2016; Smith et al., 2020; Iyer and Das., 2021). In this context Cdu1 functions as an atypical metaeffector since it targets multiple effectors and unlike other effector-metaeffector pairs it is not encoded in the same locus as its targets. We also observed that Cdu1 interactions with InaC, IpaM, and CTL0480 likely occur independently from each other and that the stoichiometry and kinetics of degradation varies for each Inc during infection of host cells. For example, InaC protein levels decrease as early as 24 hpi while IpaM and CTL0480 protein levels decrease at later stages of infection (36 and 48 hpi) (Figure 3). Furthermore, in the absence of Cdu1, we observe almost complete loss of InaC at 36 hpi (Figure 4A) and of CTL0480 at 48 hpi (Figures 3C and 3D), while only a subpopulation of IpaM is destabilized throughout infection (Figures 3B and 4B).

The observation that the DUB activity of Cdu1 was not required to protect InaC, IpaM, and CTL0480 from ubiquitination was unexpected and pointed to a potential role for its acetylation activity instead. Because Cdu1 can interact with InaC, IpaM, and CTL0480 we hypothesized that Cdu1 may protect these Incs through acetylation of lysine residues that are otherwise targeted for ubiquitination. Indeed, we found that all three Incs and Cdu1 are acetylated at lysines in infected cells, however, we were unable to determine if lysine acetylation in all four proteins was dependent on Cdu1s’ Act activity or if these PTMs are protective. Why the DUB activity of Cdu1 is unable to compensate for loss of its Act activity remains unknown. It is possible that Cdu1, like other DUBs, is regulated by PTMs (Komander et al., 2009a). For instance, phosphorylation of human CYLD inhibits its DUB activity towards TRAF2 while phosphorylation of human USP8 inhibits its DUB activity toward EGFR (Reiley et al., 2005; Mizuno et al., 2007). Mass spectrometry analysis of immunoprecipitated Flag tagged Cdu1 expressed in Ct revealed that Cdu1 is phosphorylated at multiple serine and threonine residues (Figure 4H) as previously suggested (Zadora et al., 2019). We identified three PX(S/T)P MAPK phosphorylation consensus sequence motifs in the proline rich domain (PRD) of Cdu1, suggesting that MAPKs may regulate the DUB activity of Cdu1.

Cdu1 homologs are found in multiple Chlamydia species including C. trachomatis, C. muridarum, C. suis, C. psitacci, C. abortus, C. caviae, and C. felis but is notably absent in the genomes of C. pneumoniae and C. pecorum. The acquisition of a second deubiquitinase paralog (Cdu2) has also occurred in C. trachomatis, C. muridarum, and C. suis. In the genomes of all three species, cdu2 resides directly adjacent to cdu1; an arrangement that presumptively arose from a gene duplication event. Cdu2 is a dedicated ULP with deubiquitinating and deneddylating activities (Misaghi et al., 2006; Pruneda et al., 2016). New evidence suggest that both paralogues might not be functionally redundant. The crystal structure of Cdu2 has revealed differences in residues involved in substrate recognition between Cdu1 and Cdu2 and that each paralog might recognize polyUb chains differently (Hausman et al., 2020). The processivity rates for removal of terminal Ub from polyUb chains also differs between both isopeptidases with Cdu2 exhibiting limited trimming of polyUb as compared to Cdu1 (Hausman et al., 2020). Moreover, Cdu2 lacks the proline rich domain found in Cdu1 which might be important for regulation of Cdu1 enzymatic activity. The presence of Cdu2 might also explain the low incidence of human and Ct proteins that were differentially ubiquitinated in the absence of Cdu1 (Figure 1). Whereas several Chlamydia species have acquired either one or two deubiquitinase paralogs, both C. pneumoniae and C. pecorum have not. Instead, both species have acquired an unrelated deubiquitinase (ChlaOTU) belonging to the OTU family of proteases (Makarova et al., 2000; Furtado et al., 2013). Curiously, ChlaOTU is also found in C. psitacci, C. abortus, C. caviae, and C. felis all of which encode only Cdu1 and is absent in C. trachomatis, C. muridarum, and C. suis, all of which encode Cdu1 and Cdu2. It is noteworthy that Chlamydia species have independently acquired deubiquitinases multiple times (Cdu1, Cdu2, ChlaOTU) and that some of these deubiquitinases have evolved into moonlighting enzymes reflecting the diverse strategies adopted by pathogenic Chlamydia as they adapt to their particular niche.

Supporting information

eLife Assessment

This important study combines state-of-the art proteomics and genetic manipulation of Chlamydia trachomatis to study the function of a chlamydial effector, Cdu1, with deubiquitination and acetylation activities. Solid evidence is provided to show that Cdu1 is able to protect itself and three other chlamydial effectors, which are involved in the control of chlamydial egress from host cells, from ubiquitin-mediated degradation, and that this depends on the acetylation activity of Cdu1, but not on its deubiquitination activity. This work should be of interest to microbiologists and cell biologists studying host cell-pathogen interactions.

Reviewer #1 (Public Review):

The initial goal of this work was to study how the activity of the C. trachomatis effector Cdu1 impacts on the number and nature of ubiquitinated proteins in infected host cells, and how this is related to a previously described function of Cdu1 in promoting Golgi distribution around the Chlamydia vacuole, known as inclusion.

The authors generated a cdu1-null mutant in C. trachomatis and used proteomics to analyse ubiquitinated proteins in cells infected with Cdu1-producing and Cdu1-deficient chlamydiae, by comparison to mock-infected cells. It was found that among the four proteins specifically ubiquitinated after infection with Cdu1-deficient chlamydiae there were three other C. trachomatis effectors (InaC, IpaM and CTL0480). These three proteins are part of a large family of Chlamydia effectors, known as Incs, that insert in the inclusion membrane.

Based on these observations, the authors then focused in understanding how Cdu1 protects InaC, IpaM and CTL0480 from ubiquitination, and what are the consequences of this protection for the protein levels of these Incs and for their functions during infection. It is shown that Cdu1 can bind InaC, IpaM and CTL0480, and protects these Incs and itself from ubiquitination and proteasomal degradation. This protective function of Cdu1 depends on its acetylation, but not on its deubiquitinating activity, and host cells infected by the cdu1 null mutant show defects that phenocopy those of cells infected by inaC, ipaM or ctl0480 null-mutants.

Finally, it was previously shown that CLT0480 controls/inhibits a pathway of chlamydial egress from host cells involving extrusion of the entire inclusion. The authors show that InaC and IpaM also control/promote extrusion of C. trachomatis inclusion and that the cdu1 null mutant also shows a defect in this process. This leads to the conclusion stated in the title that Cdu1 regulates chlamydial exit from host cells by protecting specific C. trachomatis effectors from degradation.

This is an excellent and impressive work, both from technical and conceptual perspectives, which accomplishes the goal of providing mechanistic insights on the mode of action of Cdu1. Overall, the data provides solid evidence for the proposed model by which the acetylation activity of Cdu1 protects itself and three Incs (InaC, IpaM and CTL0480) from degradation.

I agree that (all together) the data provides a solid support for the idea that the multiple phenotypes displayed by cells infected with the cdu1 null mutant are related to the decreased levels of InaC, IpaM and CTL0480. However, to some extent, these Incs can still be detected in cells infected with the cdu1 null mutant and it cannot be formally excluded that Cdu1 directly promotes assembly of F-actin and Golgi repositioning around the inclusion, MYPT1 recruitment to the inclusion, and extrusion of the inclusion.

Still, I think the major significance of this work comes from the combined use of proteomics and chlamydial genetics to disclose a unique a mechanism by which one effector controls the levels of other effectors. This further emphasizes that for a single bacterium injecting dozens of effectors into host cells, the function of one bacterial effector can control, and be controlled by other effectors.

Reviewer #2 (Public Review):

The manuscript describes the detailed characterization of the C. trachomatis protein Cdu1. Previous work that laid the foundation identified two enzymatic activities associated with Cdu1 - deubiquitinase and transacetylase. This work advances current knowledge by identifying Cdu1 targets for stabilization, and establishing the relationship between the two activities of Cdu1. Furthermore, the authors determined that Cdu1 is subject to autostabilization. In addition to the novelty of the findings, the strength of this report is its scientific rigor, with several experimental evidence independently confirmed using a variety of approaches, including the creation of mutants that decoupled deubiquitination from transacetylase activity. Another strength is the direct demonstration of transacetylation of the targets, which increased the relevance of the reported colocalization and interaction of Cdu1 with the targets.

The authors also made a convincing case for the basis of Cdu1 modification of each of the effector targets by linking loss of acetylation with decreased stability. An unexpected result, at least to this reviewer is the requirement for the three effectors in chlamydial egress by extrusion of the inclusion. Cdu1 regulating all three effectors underscores the importance of the timing and efficiency of inclusion extrusion. Additional insights into how the three effectors interact functionally could be obtained by specifically monitoring the timing of extrusion. Data for CTL0480 points to a negative regulator of extrusion, which could be at the level of timing, in addition to efficiency.

Overall, the work is rigorous, and makes important contribution to our understanding of the significance of Cdu1 function in in vitro infection.

Reviewer #3 (Public Review):

In this article by Bastidas et al. the authors examine the functions of the Chlamydia deubiquitinating enzyme 1 (Cdu1) during infections of human cells. First, a mutant lacking Cdu1 but not Cdu2 was constructed using targetron and quantitative proteomics was used to identify differences in ubiquitinated proteins (both host and bacterial) during infection. While they found minimal changes in host protein ubiquitination, they identified three Chlamydia effector proteins, IpaM, InaC and CTL0480 were all ubiquitinated in the absence of Cdu1. Microscopy and immunoprecipitations found Cdu1 directly interacts with these Chlamydia effectors and confirmed that Cdu1 mediates the stabilization of these effectors at the inclusion membrane during late infection time points. Surprisingly rather than deubiquitination driving this stabilization, the acetylation function of Cdu1 was required, and acetylation on lysine residues prevented degradative ubiquitination of Cdu1, IpaM, InaC and CTL0480. In line with this observation the authors show that loss of Cdu1 phenocopies the loss of single effector mutants of InaC, IpaM and CTL0480, including golgi stack formation and the recruitment of MYPT1 to the inclusion. The aggregation of changes to the Chlamydia inclusion does not alter growth but controls extrusion of chlamydia from cells with reduced extrusion in Cdu1 mutant Chlamydia infections. The strengths of the manuscript are the range of assays used to convincingly examine the biochemical and cellular biology underlying Cdu1 functions. The finding that acetylation of lysine residues is a mechanisms for bacterial effectors to block degradative ubiqutination is impactful and will open new investigations into this mechanism for many intracellular pathogens. There are a few weaknesses that temper enthusiasm for the manuscript in its current form. These include caveats related to the timing of proteomics, the lack of an effect of Cdu1 directly on bacterial growth, and discussion of previous studies. Altogether this is an important series of findings that help to understand the mechanisms underpinning Chlamydia pathogenesis using orthologous methods with a few caveats that lower the overall impact.

Author Response

The following is the authors’ response to the original reviews.

RESPONSE TO REVIEWERS:

Reviewer #1 (Recommendations For The Authors):

I think the manuscript of this excellent work can be improved, especially in writing (including a suggestion in the title) and presentation (Figure 6); Also some additional specific experiments and analyses could be important, as I suggest below,

1. For the title, perhaps a shorter "The acetylase activity of Cdu1 protects Chlamydia effectors from degradation" would be better to convey the major significance of this work. Of course, Cdu1 must regulate the function of InaC, IpaM and CTL0480. But perhaps it is speculative to think that egress is the major function of these effectors as their activity on other host cell processes during the cycle could eventually impact the extrusion process indirectly.

Although we concur with the insights provided by reviewer 1, we wish to underscore that a significant breakthrough presented in our study revolves around the regulation of Chlamydia exit by Cdu1. Consequently, we believe that this noteworthy discovery should be incorporated into the title.

1. For the writing:

a. The description of ubiquitination and DUBs could be synthesized to the essential, so that space is gained to explain things that then come a bit out of the blue in the results (what are Incs, the specific functions of InaC, IpaM, and CTL0480 - at least place the citations in lines 110-112 next to the corresponding Incs -, Cdu2, etc - see specifics below)

In lines 182-196 of the revised manuscript, we have incorporated additional contextual information concerning the roles of Incs, along with descriptions of the functions of InaC, IpaM, and CTL0480.

b. In the Results, there is a lot of Chlamydia- and maybe lab-specific jargon that could be significantly simplified for the more general reader. I detail some suggestions below in the specific issues.

We have improved the readability of our manuscript for a general audience by removing Chlamydia-specific terminology from the entire text and figures.

1. For the figures:

a. Figure 6, this figure could be reorganized: why two graphs in panel D? If detailed quantifications were done, perhaps in panel B just zoom on the examples of Golgi distributed/compacted? And again the labelling Rif-R L2, L2 pBOMB, M407 p2TK2, etc, simplify?

Figure 6 has undergone restructuring. The representative images have been relocated to Supplemental Figures 5 and 6, while we have introduced sample images demonstrating F-actin assembly and Golgi repositioning. Furthermore, the quantification of Golgi dispersal has been streamlined into a single panel. Additionally, we have simplified the labeling of the strains utilized in the study.

b. Figure 3, in the labelling, WT, inaC null, cdu1::GII wouldn't be enough? Leave the details to the legend and/or M&M.

We have simplified the labeling of Ct strains in Figure 3.

c. Figure 3C, these arrowheads should not be so symmetric (small arrows instead?) and it is unclear that the indicated cells do not show CTL0480.

We have substituted arrowheads with small arrow symbols and have also revised the Figure to incorporate a new representative image that prominently illustrates the absence of CTL0480 at the inclusion membrane of some cdu1::GII inclusions within infected Hela cells at 36 hpi.

1. Experiments:

a. In Figure 7, at least extrusion should be analysed also with the Cdu1-deficient strain expressing Ac-deficient Cdu1 and the inaC and ipaM phenotypes should be complemented.

We have conducted additional experiments to analyze extrusion production in Hela cells infected with a cdu1 null strain expressing the acetylase-deficient Cdu1 variant. We have incorporated the relevant data into revised Figure 7, where the impact of this strain on extrusion production and size is presented. Additionally, we updated Supplemental Figure 8 to include data illustrating the number of inclusions produced by this strain. We have also addressed these new results in the revised manuscript (lines 424-432). We are currently complementing inaC and ipaM mutant strains with various InaC and IpaM constructs that will be used in a follow up manuscript.

b. Does overexpression of InaC, IpaM, or CTL0480 in a cdu1-null background prevent the degradation of these Incs and suppress the defects of cells infected by the cdu1 mutant (F-actin, Golgi, MYPT1)? This would show that the multiple phenotypes displayed by cells infected by the cdu1 null mutant are indeed related to the decreased levels of InaC, IpaM and CTL0480.

We opted not to include data from the overexpression of these effectors in a cdu1-null background due to an unexpected decrease in shuttle plasmid load during overexpression. This development prompted concerns regarding the potential detrimental effects of overexpressing these effectors in the absence of Cdu1. Data supporting this observation are not included in this report.

c. Figures 3A and 3B should be quantified (it says it is from 3 independent experiments). It would be important to have a relative perspective of how much Cdu1 protects these Incs over time (for InaC, it would also be nice to have the 36 and 48 hpi time-point). This is in contrast with the microscopy data in Figure 5, which illustrates very clear effects, and the quantification is a bit redundant.

In Figure 3, we have incorporated a new Western Blot image showing endogenous InaC protein levels in Hela cells following infection with both WT Ct and cdu1::GII strains at 24, 36, and 48 hours post-infection (hpi). Additionally, we have quantified the Western Blot signals for both InaC and IpaM, and these results are also presented in Figure 3. The quantification of MYPT1 recruitment has been relocated to a supplementary figure. We have also included details regarding the methodology employed for the quantification of Western Blot signals in the Materials and Methods section.

d. What is the subcellular localization of InaC, IpaM, CTL0480 and Cdu1 when analysed by transfection? Does Cdu1 bind to of InaC, IpaM, CTL0480 in infected cells? If this was attempted and unsuccessful it should be mentioned.

In transfected HEK cells, InaC, IpaM, CTL0480, and Cdu1 all exhibit cytoplasmic localization with a diffuse pattern (data not shown). Despite our efforts, we encountered challenges in observing co-immunoprecipitation of Cdu1 with all three Incs in infected Hela cells at 24 hpi, We have duly acknowledged this limitation in our findings, as reflected in line 221-226 of the revised manuscript.

1. Specific issues:

Line 87, "propagule" is really needed to describe the EB?

The EB is the infectious form of Chlamydia species that spreads within the host to renew its life cycle; thus, "propagule" is a suitable term to characterize the EB.

Exocytosis implies fusion with the plasma membrane so "inclusion is exocytosed" (line 91) is not entirely correct.

In line 91 of the revised manuscript, we referred to extrusion as the exit of an intact inclusion from the host cell and omitted the use of "exocytosed" to describe this process.

Line 126, "a Ct L2 (LGV L2 434 Bu) background". Maybe "a Ct cdu1-null strain" would be enough and leave the detail for Materials and Methods.

In line 128 of the revised manuscript, we omitted "(LGV L2 434 Bu)" to avoid using jargon that may be unfamiliar to readers not well-versed in Chlamydia terminology.

Line 138, in the previous Pruneda et al, Nature Microbiol 2018, the title of figure 4 is "ChlaDUB deubiquitinase activity is required for C. trachomatis Golgi fragmentation", so why raise this hypothesis? And why in the end is the acetylation activity of Cdu1 that promotes Golgi distribution? I think this related with infection vs transfection experiments but it deserved to be briefly explained/discussed.

In lines 140-142 of the revised manuscript, we provide clarification that the DUB activity of Cdu1 is required for Golgi fragmentation in transfected cells. This observation supports our initial hypothesis suggesting that the DUB activity of Cdu1 is also required for Golgi distribution in infected cells, and our rationale for identifying targets of its DUB activity.

Lines 147-155, what is the relevance of this non-ubiquitinated proteins that come along? Couldn't this be synthesized?

We have included a discussion on non-ubiquitinated proteins, as they could potentially encompass proteins that interact with those protected by Cdu1. This perspective provides supplementary insights into the roles of proteins targeted for ubiquitination in the absence of Cdu1. The results of this analysis have been succinctly summarized in a single paragraph within the initial manuscript (lines 151-159 of the revised manuscript).

Line 170, I think it is the first time that "Type 3 secretion"; perhaps explain in the introduction.

Type 3 secretion systems have been extensively characterized and discussed in the literature, and we anticipate that the majority of our readers are well-acquainted with this secretory mechanism.

Line 184, I think it is the first time "microdomains" are mentioned; perhaps mention in the introduction.

The definition of "microdomains" has been provided in line 191 of the revised manuscript.

Figure 2, as it stands the analysis with truncated Cdu1 proteins adds little to the work. Binding to the Incs seems to be affected when the TM domain is not present, but it still binds. And this is in a transfection context.

The results depicted in Figure 2, involving truncated Cdu1 proteins, illustrates that Cdu1 is capable of interacting with InaC, IpaM, and CTL0480 even in the absence of infection. This finding serves as evidence suggesting that all three Incs could potentially serve as direct targets for Cdu1 activity. As a result, we prefer to keep these findings in the manuscript.

Line 219, "late stages of infection", this is shown (albeit not completely quantified) for IpaM and CTL0480, but not for InaC.

In the revised Figure 3, we show InaC protein levels at 24, 36, and 48 hours post-infection, and we have incorporated quantitative data for both InaC and IpaM protein levels in the context of Hela cells infected with both WT L2 and cdu1::GII strains. This updated figure serves to emphasize the pivotal role of Cdu1 in safeguarding all three Incs during the late stages of infection.

Line 233, "pBOMB-MCI backbone" - is this needed in the Results section? And this refers to Figure 4 while pBOMB appear already in Fig. 3.

We have removed “pBOMB-MCI backbone” in the revised manuscript.

Line 236, should be cdu1 endogenous promoter.

In line 265 of the revised manuscript we have replaced Cdu1 with cdu1 (italicized).

Line 263, WT.

In line 293 of the revised manuscript we replaced “wild type” with “WT”.

Line 277, IncA instead of "the Inc protein IncA".

In the manuscript we wanted to emphasize that IncA is also an inclusion membrane protein, therefore we have included “the Inc protein IncA” in the revised manuscript to avoid any confusion.

How does the data in Figure 5 relates to the relatively few proteins ubiquitinated in cells infected with cdu1-mutant Ct? These Ub-labelling corresponds to ubiquitinated InaC, IpaM and CTL0480?

The findings presented in Figure 5 demonstrate that the acetylase activity of Cdu1 plays a crucial role in enabling Ct to block all ubiquitination events taking place on or in proximity to the periphery of the inclusion membrane. This encompasses Cdu1 targets that might not have been identified through our proteomic analysis.

Lines 299-301, "M923 inclusions", there is certainly a clear way to write this.

In lines 326-327 and 332-332 of the revised manuscript, we have clarified that “M923” is an incA null strain to provide clarification.

Line 309, is "peripheries" correct?

We have changed “peripheries” with “periphery” in the revised manuscript (line 360).

Line 312, "Rif-R L2" and "M407" - can this be simplified?

In the revised manuscript, "Rif-R L2" was substituted with "WT L2" in lines 363 and 382, while "M407" was exchanged with "an inaC null strain" in lines 311, 367, and 368. These same replacements were applied to the Figures and their corresponding legends for consistency.

Lines 308-321, and 326-335, these % are all approximate figures and this should be made clear.

In lines 364-395 of the revised manuscript we have stated that all percentages are approximate values.

Fig. S1, kb and not k.b; what's the "+ control"; and is not really possible to have a PCR that works for the *? 3 kb is not that long.

In the updated Figure S1, we have corrected "k.b" to "kb". In the legend of Figure S1, we have clarified that the + control corresponds to the cdu2 locus. Moreover, we could not cleanly amplify a 3 kb PCR product from bacteria in whole cell lysates of infected mammalian cells (Vero cells).

Fig. S2, kb and not k.b, bp and not b.p

In the updated Figure S2, we have corrected “k.b” with “kb” and “b.p” with “bp”.

Reviewer #2 (Recommendations For The Authors):

Figure 1 describes an affinity-based purification and mass spectrometric identification of differentially ubiquitinated proteins (host and chlamydial). Through different permutations of combinations of infection (mock, wild type, and Cdu1 mutant), three effectors, IpaM, InaC, and CTL0480, were identified as putative targets of Cdu1. The authors used a high-stringency cutoff, which could explain identification of only three targets. Having said this, the localization of Cdu1 to the inclusion membrane would be expected to also narrow down the number of targets. Interestingly, Cdu2, another deubiquitinase remained active in these experiments, which could have affected identification of Cdu1 targets. The authors addressed this issue by referring to previously reported structural studies. A somewhat glaring omission is the lack of reference to NF-kB as a substrate of ChlaDub1/Cdu1. In experiments by Le Negrate et al., ChlaDub1 ectopic overexpression in cells led to the deubiquitination of IkB-alpha, thus inhibiting the nuclear translation of NF-kB. Based on the inclusion membrane localization of Cdu1 during infection, is the identification of IkB an artifact of overexpression of Cdu1, or is it still a bona fide Cdu1 target?

We conducted experiments using our cdu1 null strain to investigate whether IκBα could be a target of Cdu1 activity. While our findings are intriguing and relevant, it is not feasible to determine, at this stage, whether our findings result from a direct or indirect consequence of Cdu1 localizing to the inclusion membrane. Consequently, these findings extend beyond the scope of the current manuscript. We plan to explore the implications of our observations more deeply in a subsequent manuscript, where we intend to provide a more comprehensive and mechanistic analysis based on these preliminary findings. Additionally, we have referenced the potential targeting of IκBα by Cdu1 in lines 100-101 and 166-171 of the revised manuscript.

Figure 2 demonstrates the individual interaction of the identified effectors with Cdu1. Interaction at the inclusion membrane is inferred from colocalization studies, while protein-protein interaction is monitored using ectopic overexpression of tagged versions of Cdu1 and the individual effectors. This is somewhat of a weakness of the manuscript because the mechanism of action of Cdu1 towards its target hinges on protein-protein interaction.

Despite our efforts, we encountered challenges in co-immunoprecipitating endogenous Cdu1 with all three Incs in infected Hela cells at 24 hpi. There are multiple technical reasons as to why these interactions, which are predicted to be transient, will not be captured by bulk affinity approaches such as immunoprecipitations, especially when the starting materials are present in very low abundance. We acknowledged these limitations in our findings, as reflected in lines 221-226 of the revised manuscript.

Figure 3 provides the first evidence in this paper of the importance of the inferred interaction of Cdu1 with the three effectors. The authors show that the loss of cdu1 has stability consequences on the three effectors. This figure would benefit from quantifying InaC- or IpaM-positive inclusions in the same manner done with CTL0480. The timepoint-dependent effect of Cdu1 loss of function is intriguing. Do InaC and IpaM retention at the inclusion show the same timepoint-dependent characteristic?

In the revised Figure 3, we have incorporated InaC protein levels at 24, 36, and 48 hours post-infection. Additionally, we have included quantitative data representing both InaC and IpaM protein levels in HeLa cells infected with both WT L2 and cdu1::GII strains. The quantification of CTL0480 localization to cdu1::GII inclusions has been moved to a supplementary figure.

This updated figure illustrates that the absence of Cdu1 has a time-dependent impact on both InaC and IpaM. However, it is noteworthy that the kinetics of degradation for these two proteins diverge significantly.

For Figure 7, the authors should consider monitoring timing of inclusion extrusion to gain additional insight into the functional interactions between the effectors. For example, the loss of CTL0480 leads to increased extrusion, implying a role in delaying or suppressing extrusion. In a time-course experiment, a CTL0480 mutant could exhibit an earlier occurrence of inclusion extrusion.

One of the principal discoveries of this study is that Cdu1, InaC, IpaM, and CTL0480 collaborate to facilitate optimal extrusion of Ct from host cells. These findings represent a significant contribution to our understanding of how Chlamydia controls its exit from infected cells. We are currently in the process of expanding on these results. A forthcoming follow-up manuscript will provide more detailed and comprehensive exploration of these findings.

Reviewer #3 (Recommendations For The Authors):

Specific comments.

a. I have some concerns related to the time point chosen for mass spec analysis and potential caveats and alternative interpretations. This work was done relatively early (24 hours) compared to the most convincing Cdu1 functions that occur later, thus this may limit the authors global understanding of protein changes. For example, the known substrate of Cdu1, Mcl-1 was not identified but this is altered relatively late during infection. Thus, the surprise that minimal host proteins are altered in ubiquitination may be partially driven by the timing of the assay. This should be more clearly discussed as a caveat.

In the revised manuscript (lines 166-171), we have acknowledged that there might be additional targets of Cdu1 that remain unidentified, primarily due to the specific time point we utilized in our study.

b. Another caveat to these studies is while the loss of Cdu1 alters different effectors stability and function and extrusion size, these changes do not modulate bacterial growth in cells. The authors speculate that regulating extrusion size may alter interactions with innate cells to drive dissemination. However, a previous study found defects in an animal model using a Cdu1 transposon mutant found decreased bacterial load in the genital tract. It is also possible that redundancy of effectors may mask importance in growth of Cdu1, but the authors strongly argue against redundancy of Cdu1 and Cdu2 so this weakens the authors argument here. These concepts and published data should be more directly discussed in the context of the authors proposed extrusion model and the role in driving Chlamydia growth and pathogenesis.

In our revised manuscript (lines 460-466) we propose that while we do not observe any growth impairments during Ct growth in the absence of Cdu1 in HeLa cells, the reduction in bacterial loads observed in murine models of infection with an independent cdu1 mutant strain (cdu1::Tn) may potentially be linked to defects in extrusion production or alterations in Cdu1-dependent regulation of extrusion size.

c. Recent studies have found that IFNg activation can result in dramatic changes in ubiquitination to pathogen containing vacuoles. While some of these are blocked by the newly found GarD, it seems possible that Cdu1 may also play a role (and perhaps use its deubiquinating activity) to further protect the inclusion. In light of published results showing that Cdu1 mutants have lower IFU burst size only in IFNg activated cells, this may be an important caveat in the current studies. This should be more directly addressed in the current manuscript.

We have incorporated two experimental findings indicating that the presence of Cdu1 is not required for Ct to defend itself against IFN cellular immunity in human cells. These recent discoveries are now presented in the updated Figure 5 and detailed in lines 338-355 of the revised manuscript.

d. On lines 433-434 the authors claim that Cdu1 is atypical since it is not encoded with the metaeffector/target pairs. However, this is an oversimplification of what is known about metaeffectors. For example, there are meta-effector/effector pairs that are not encoded together in Legionella (see table 1 DOI: https://doi.org/10.3390/pathogens10020108). Thus, the discussion should be adjusted. It seems Cdu1 is the first meta-effector found in Chlamydia, and maybe this should be highlighted more strongly rather than its uniqueness in this aspect of meta-effector/effector functions.

In lines 488-489 of the revised manuscript, we have removed the assertion that Cdu1 functions as an atypical metaeffector and emphasized that it represents the initial discovery of a metaeffector within Ct.

Version 2

Abstract

Abstract

Many cellular processes are regulated by ubiquitin-mediated proteasomal degradation. Pathogens can regulate eukaryotic proteolysis through the delivery of proteins with de-ubiquitinating (DUB) activities. The obligate intracellular pathogen Chlamydia trachomatis secretes Cdu1 (ChlaDUB1), a dual deubiquitinase and Lys-acetyltransferase, that promotes Golgi remodeling and survival of infected host cells presumably by regulating the ubiquitination of host and bacterial proteins. Here we determined that Cdu1’s acetylase but not its DUB activity is important to protect Cdu1 from ubiquitin-mediated degradation. We further identified three C. trachomatis proteins on the pathogen-containing vacuole (InaC, IpaM, and CTL0480) that required Cdu1‘s acetylase activity for protection from degradation and determined that Cdu1 and these Cdu1-protected proteins are required for optimal egress of Chlamydia from host cells. These findings highlight a non-canonical mechanism of pathogen-mediated protection of virulence factors from degradation after their delivery into host cells and the coordinated regulation of secreted effector proteins.

Introduction

Ubiquitination is a conserved and ubiquitous post-translational modification (PTM) of proteins involving the conjugation of the carboxy-terminal glycine residue of ubiquitin (Ub) to lysine residues of target proteins. Poly-ubiquitination of substrates involves further conjugation of a Ub internal lysine residue or amino-terminal methionine (M1) with a second Ub molecule. Seven internal lysines in Ub (K6, K11, K27, K29, K33, K48, K63) and M1 are utilized by Ub conjugating enzymes to form homogeneous, branched, or mixed poly-ubiquitin (polyUb) chains (Komander and Rape., 2012). PolyUb chains with different linkage types exhibit distinct structures and functions. For example, K48- and K11-linked polyUb chains exhibit a compact conformation and are substrates for 26S proteasome-mediated degradation (Varadan et al., 2004; Tenno et al., 2004; Eddins et al., 2007; Bremm et al., 2010; Saeki., 2017). In contrast, K63-linked polyUb conjugates adopt more open conformations that enable the recruitment of multiprotein complexes that regulate the function of the target protein by proteolytic independent events (Komander et al., 2009b; Weeks et al., 2009; Datta et al., 2009; Komander and Rape., 2012). Mixed and branched polyUb chains are also emerging as important regulators of physiological functions (Swatek and Komander., 2016; Ohtake and Tsuchiya., 2017).

Protein ubiquitination regulates numerous eukaryotic cell processes including protein degradation, signal transduction, cell cycle regulation, selective autophagy, the DNA damage response, and programmed cell death. Ub also plays key roles in modulating host innate immune responses to bacterial infection (Li et al., 2016), bacterial proteins, pathogen-containing vacuoles, and bacteria themselves by targeting them for Ub-mediated degradation by proteasomal or autophagic machineries (Li et al., 2016). Because the Ub system is critical for pathogen containment, many pathogens have evolved mechanisms to counteract the impact of this PTM (Vozandychova et al., 2021). For instance, bacterial deubiquitinases (DUBs) can remove Ub from ubiquitinated substrates thereby dampening inflammatory and cell-autonomous defense mechanisms (Kubori et al., 2019). Many DUBs are cysteine proteases with a catalytic Cys, a nearby His and an Asn/Asp (Komander et al., 2009a). DUBs are typically dedicated to the removal of Ub moieties and are unable to hydrolyze other Ub-like (Ubl) modifications such as SUMO or NEDD8. However, the CE clan of Ubl proteases (ULPs) can catalyze the removal of both SUMO and NEDD8 (Ronau et al., 2016). Bacterial pathogens also encode CE clan enzymes that function as DUBs, ULPs or both. For instance, Salmonella Typhimurium SseL, Escherichia coli ElaD, and Shigella flexneri ShiCE function as Ub specific proteases (Rytkönen et al., 2007; Catic et al., 2007; Pruneda et al., 2016) while RickCE from Rickettsia belli functions as a protease directed towards both Ub and NEDD8 as does SidE from Legionella pneumophila which displays mixed activities towards Ub, NEDD8, and ISG15 (Sheedlo et al., 2015; Pruneda et al., 2016). Similarly, XopD from Xanthamonas campestris and LotB from L. pneumophila are isopeptidases exhibiting cross reactivity towards both Ub and SUMO (Pruneda et al., 2016; Schubert et al., 2020). Some CE clan bacterial effectors display acetyltransferase activity. L. pneumophila LegCE, S. Typhimurium AvrA, and YopJ from Yersinia pestis function exclusively as acetyltransferases (Mittal et al., 2006; Mukherjee et al., 2006; Jones et al., 2008; Pruneda et al., 2016). In contrast, the Chlamydia trachomatis (Ct) effector Cdu1/ChlaDUB1 is a CE clan protein that exhibits both acetyltransferase and deubiquitinating activities (Misaghi et al., 2006; Pruneda et al., 2016; Fischer et al., 2017; Pruneda et al., 2018).

Ct is an obligate intracellular bacterial pathogen responsible for human diseases of significant clinical and public health importance (Haggerty et al., 2010). Ct has a biphasic developmental cycle in which the Ct infectious propagule or elementary body (EB) invades the target host cell. Upon internalization the EB transitions to the reticulate body (RB). RBs replicate by binary fission within a pathogenic vacuole (“inclusion”) and asynchronously differentiate back to EBs. In cell culture, starting at around 48 hours-post infection (hpi) Ct will egress after lysis of the host cell or by a process termed extrusion, wherein the intact inclusion exits from the infected cell (Moulder., 1991; Abdelrahman and Belland., 2005; Hybiske and Stephens., 2007; Lee et al., 2018). The effector Cdu1 was originally identified as a deneddylating and deubiquitinating enzyme and subsequently shown to exhibit in-vitro isopeptidase activity towards both Lys48 and Lys63 linked di-Ub substrates (Misaghi et al., 2006; Claessen et al., 2013; Pruneda et al., 2016; Fischer et al., 2017). Cdu1 is unique among CE clan enzymes in that it also functions as a bona fide lysine acetylase with both is acetylase and DUB activities catalyzed by the same catalytic active site (Pruneda et al., 2018). Intriguingly, Cdu1 autoacetylation is directed towards lysines unlike other CE clan acetylases that predominantly target serine and threonine residues (Pruneda et al., 2018). In transfected cells, Cdu1 protects the NFκB cytoplasmic retention factor IκBα from ubiquitination and proteasomal degradation (Le Negrate et al., 2008). In infected cells, Cdu1 localizes to the inclusion membrane where it functions to stabilize the anti-apoptotic protein Mcl-1 and to promote the repositioning of Golgi ministacks around the Ct inclusion (Fischer et al., 2017; Wang et al., 2018; Pruneda et al., 2018; Kunz et al., 2019; Auer et al., 2020). However, the mechanism by which Cdu1 promotes redeployment of Golgi ministacks and any additional roles that Cdu1 may play during Ct infection of epithelial cells remains unknown.

In this study, we show that Cdu1 protects itself and three secreted Ct effectors, InaC, IpaM, and CTL0480 from targeted ubiquitination and proteasomal degradation. InaC, IpaM, and CTL0480 are members of a larger family of bacterial proteins embedded within the inclusion membrane (Inc proteins) (Bannantine et al., 2000; Rockey et al., 2002; Chen et al., 2006; Li et al., 2008; Alzhanov et al., 2009; Dehoux et al., 2011; Lutter et al., 2012; Lutter et al., 2013; Kokes et al., 2015; Weber et al., 2015). We show that Cdu1-mediated protection from degradation is independent from its DUB activity but relies upon its Lys acetylase activity. We show that Cdu1 protects InaC to promote repositioning of Golgi ministacks and formation of actin scaffolds around the Ct inclusion, and CTL0480 to promote recruitment of myosin phosphatase target subunit 1 (MYPT1) to the inclusion. In addition, we determined that Cdu1 and Cdu1-protected Incs are required for optimal extrusion of inclusions from host cells at the late stages of infection.

Results

The C. trachomatis inclusion membrane proteins InaC, IpaM, and CTL0480 are differentially ubiquitinated in the absence of Cdu1

Cdu1 is required for Golgi repositioning around the Ct inclusion (Pruneda et al., 2018; Auer et al., 2020). To understand how Cdu1 promotes Golgi redistribution, we first generated a cdu1 null strain in a Ct L2 background by TargeTron mediated insertional mutagenesis (pDFTT3-aadA) (Lowden et al., 2015) (S. Figure 1). Loss of Cdu1 expression in the resulting L2 cdu1::GII aadA (cdu1::GII) strain was verified by western blot analysis and by indirect immunofluorescence with antibodies raised against Cdu1 (S. Figures 2A and 2B). Because cdu2 resides directly downstream of the cdu1 locus and encodes a second Ct DUB (Cdu2/ChlaDUB2) (Misaghi et al., 2006), we first determined whether the disruption of cdu1 impacted the expression of cdu2. We detected cdu1 and cdu2 transcripts in HeLa cells infected with Ct L2 but not for the juncture between cdu1 and cdu2 (S. Figure 2C). In cells infected with Ct cdu1::GII we only detected cdu2 transcripts (S. Figure 2C) confirming that cdu1 and cdu2 are not co-expressed as part of an operon in accordance with previous observations (Albrecht et al., 2010).

In transfected Hela cells, Cdu1’s DUB activity has been linked to fragmentation of the Golgi apparatus (Pruneda et al., 2018). We therefore hypothesized that Cdu1s’ DUB activity in infected cells promoted Golgi redistribution around inclusions and that we could identify potential targets by comparing the protein ubiquitination profile of cells infected with WT or cdu1::GII strains by quantitative mass spectrometry (MS). HeLa cell were mock infected or infected with either WT L2 or cdu1::GII strains. At 24 hpi, poly-ubiquitinated proteins were enriched from lysed cells using Tandem Ubiquitin Binding Entities (TUBEs) (LifeSensors). TUBEs consist of concatenated Ub binding associated domains (UBAs) that bind to polyUb-modified proteins with nanomolar affinities. Poly-ubiquitinated proteins of both human and Ct origin were enriched and identified by quantitative LC-MS/MS analysis.

Over 2,000 non-ubiquitinated proteins co-precipitated with TUBE 1 bound proteins across all three conditions (mock, L2, and cdu1::GII infected HeLa cells) and 3 biological replicates (S. Table 1). Among these, 47 human proteins were significantly enriched in mock infected HeLa cells and 50 human proteins were significantly enriched during Ct infection (L2 and cdu1::GII) (S. Table 3, S. Figure 3) Pathway enrichment analysis revealed that proteins involved in RNA metabolism were overrepresented among co-precipitating proteins from mock infected cells (S. Table 4, S. Figure 4) while no biological pathways or processes were overrepresented in proteins enriched from infected cells (S. Figure 4). We also identified 8 TUBE1 co-precipitating Ct proteins in HeLa cells infected with L2 and cdu1::GII (S. Table 5, S. Figure 3).

TUBE 1 affinity capture lead to the identification of 43 ubiquitinated proteins (35 human proteins and 8 Ct proteins across all 3 conditions and replicates) based on the presence of peptides containing a di-glycine remnant motif (Peng et al., 2003) (S. Tables 6-8). The lack of widespread poly-ubiquitination of either human or Ct proteins in response to Ct infection (Figure 1) was surprising given that wholesale changes in protein ubiquitination has been reported during infection of HeLa cells by intracellular pathogens like S. Typhimurium (Fiskin et al., 2016). It is also possible that had we conducted our analysis at different time points post-infection (hpi), we might have identified additional Cdu1 targets, such as Mcl1 and IκBα (Le Negrate et al., 2008; Fischer et al., 2017) which were not identified in our analysis. However, given that we observed Cdu1 at the inclusion membrane as early as 1 hpi (data not shown), we opted to focus on an earlier stage of the infection cycle. Only two human ubiquitinated proteins (ZC3H7A and DDIT4) were found to be significantly enriched in response to WT L2 infection at 24 hpi (Figures 1A and 1C, S. Table 8) while only one human protein (MGC3121) was preferentially ubiquitinated in HeLa cells infected with the cdu1::GII mutant strain (Figures 1B and 1C, S. Table 8). In contrast three Ct proteins, InaC (K104, K107, and K149), IpaM (K29), and CTL0480 (K115) were ubiquitinated at Lys residues in the absence of Cdu1 (Figures 1B, 1C, and 1D, S. Table 8).

The C. trachomatis inclusion membrane proteins InaC, IpaM, and CTL0480 are ubiquitinated in the absence of Cdu1.

(A-C) Volcano plots (pairwise comparisons) of the relative abundance of human and Ct ubiquitinated peptides.(A) Mock infected HeLa cells versus HeLa cells infected with WT Ct (L2 434 Bu pBOMB) (24 hpi). (B) Mock infected HeLa cells versus HeLa cells infected with a cdu1 null strain (cdu1::GII pBOMB) (24 hpi) (C) HeLa cells infected with WT Ct (24 hpi) versus HeLa cells infected with a cdu1 null strain (24 hpi). Significance values were interpolated from 3 independent biological replicates. Ubiquitinated proteins were enriched with magnetic TUBE 1 beads (binds to polyubiquitinated proteins) and peptides identified by quantitative LC MS/MS analysis. Three Ct inclusion membrane proteins, InaC, IpaM, and CTL0480 were differentially ubiquitinated in the absence of Cdu1. (D) InaC was ubiquitinated at K104, K107, and K149, IpaM at K290, and CTL0480 at K115 in the absence of Cdu1. TMD: Transmembrane domain. Numbering corresponds to amino acids in the protein sequence of each respective inclusion membrane protein.

InaC, IpaM, and CTL0480 are Ct effector proteins that localize to the inclusion membrane (Chen et al., 2006; Alzhanov et al., 2009; Lutter et al., 2013; Kokes et al., 2015). These Type 3 secretion substrates belong to a family of over 36 inclusion membrane proteins (Incs) that contain a signature bi-lobal hydrophobic transmembrane domain (Bannantine et al., 2000; Rockey et al., 2002; Li et al., 2008; Dehoux et al., 2011; Lutter et al., 2012). Incs provide many functions important for Ct intracellular replication ranging from providing structural integrity to the inclusion membrane, to regulating membrane trafficking, to mediating interactions with host organelles and cytoskeletal structures (reviewed in Bugalhão et al., 2019). InaC facilitates the activation of the small GTPase RhoA, a crucial step for the assembly of actin scaffolds around the inclusion (Haines et al., 2021; Kumar and Valdivia., 2008). Additionally, InaC plays a pivotal role in the activation of Arf GTPases. This activation subsequently induces post-translational modifications of microtubules in close proximity to the inclusion membrane, which are essential for Ct to initiate the repositioning of Golgi ministacks around the inclusion (Wesolowski et al., 2017). IpaM exhibits localization to discrete patches in the inclusion termed microdomains (Alzhanov et al., 2009; Dumoux et al., 2015). Upon ectopic expression, IpaM induces alterations in microtubule organization (Dumoux et al., 2015). CTL0480, facilitates the recruitment of MYPT1 (myosin phosphatase target subunit 1) to the inclusion membrane. Recrutiment of MYPT1 is required for the efficient exit of Ct from host cells (Lutter et al., 2013; Shaw et al., 2018). Because Cdu1 also localizes to the inclusion membrane (Fischer et al., 2017; Wang et al., 2018; Pruneda et al., 2018; Kunz et al., 2019) we postulated that Cdu1 directly protects InaC, IpaM, and CTL0480 from ubiquitination.

Cdu1 associates with InaC, IpaM, and CTL0480

We first determined if Cdu1 co-localized with InaC, IpaM, and CTL0480 at the inclusion membrane. HeLa cells were infected for 24 hours with WT L2 or L2 expressing CTL0480-Flag from its endogenous promoter and immunostained with antibodies against Cdu1, InaC, IpaM, or the Flag epitope. Both InaC and CTL0480-Flag localized throughout the inclusion membrane while IpaM was restricted to discrete microdomains as previously reported (Alzhanov et al., 2009; Dumoux et al., 2015) (Figure 2A). Cdu1 co-localized with InaC and CTL0480-Flag and with IpaM at microdomains (Figures 2A and 2B). All four antibodies specifically recognized their corresponding antigens since immunostaining for Cdu1, InaC, IpaM or the Flag epitope was not observed in Ct strains lacking Cdu1 (cdu1::GII), InaC (M407, Kokes et al., 2015) IpaM (ipaM::GII, Meier et al., 2022), or a strain that does not express CTL0480-Flag (Figure 2A).

Cdu1 associates with InaC, IpaM, and CTL0480.

(A) Co-localization of Cdu1(magenta) with endogenous InaC (green), IpaM (green), and ectopically expressed CTL0480-Flag (green) at the Ct (L2) inclusion membrane of HeLa cells infected for 24 h. HeLa cells infected with an inaC null strain (M407), an ipaM null strain (ipaM::GII), and WT Ct (L2 434 Bu pBOMB) were used as controls for antibody specificity. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. Images are representative of multiple images captured across three independent replicates. (B) Line scan profiles of fluorescent signal intensities displayed in (A) showing co-localization of fluorescence intensities for endogenous Cdu1 with endogenous InaC and IpaM, and with CTL0480-Flag along the L2 inclusion membrane. (C) Schematic of Cdu1-GFP(C) (L2) fusion (Cdu1-GFP) and Cdu1-GFP variants used in co-transfections of HEK 293 cells. GFP: Green fluorescent protein. TMD: Transmembrane domain. PRD: Proline rich domain. CD: Catalytic domain. FL: Full length. TMD-: Cdu1-GFP variant lacking TMD domain. CD-: Cdu1-GFP variant lacking CD domain. (D-G) Western blot analysis of GFP immunoprecipitates from HEK 293 cells co-transfected with mammalian plasmids expressing: Cdu1-GFP variants and (D) truncated 3XFlag(N)-InaC (D/UW-3/CX CT813, amino acids 96-264), (E) V5(N)-IpaM (L2, full length), (F) V5(N)-CTL0480 (L2, full length), and (G) V5(N)-CpoS (L2, full length). Vec1: Empty pOPINN-GFP vector. Vec2: Empty pDEST53 vector. Vec3: Empty pcDNATM3.1/nV5-DEST vector. Western blot images are representative from two independent experiments.

We next determined if Cdu1 can interact with InaC, IpaM, and CTL0480 by co-transfecting HEK 293 cells with vectors expressing either full length Cdu1-GFP or truncated versions of Cdu1-GFP lacking transmembrane or catalytic domains (Figure 2C), and vectors expressing Flag-InaC, V5-IpaM, and V5-CTL0480. Transfected cells were lysed and Cdu1-GFP was immunoprecipitated with antibodies against GFP. Western blot analysis of the immunoprecipitates showed that Flag-InaC, V5-IpaM, and V5-CTL0480 co-precipitated with Cdu1-GFP (Figures 2D-F). Moreover, the transmembrane domain of Cdu1 was necessary for Cdu1-GFP to interact with all three Incs (Figures 2D-F). The interaction between Cdu1-GFP and the three tagged Incs was specific since we did not detect interactions between Cdu1-GFP and a V5-tagged version of the inclusion membrane protein CpoS (Sixt et al., 2017) (Figure 2G). We expected these interactions to be transient at the inclusion membrane as the engagement of Cdu1 with its target(s) should mimic that of most enzymes with their substrates. Therefore, while we could capture these complexes in co-immunoprecipitations in the context of overexpression, this assay was not sensitive enough to reliably document the formation of complexes among low abundance endogenous Ct proteins. Nevertheless, our findings from transfection experiments lead us to conclude that Cdu1 can selectively interact with all three Incs even in the absence of infection, and these interactions are facilitated by the transmembrane domain of Cdu1.

Cdu1 protects InaC, IpaM, and CTL0480 proteins from degradation during infection

We next assessed if Cdu1 was required to stabilize endogenous InaC, IpaM, and CTL0480 in infected cells. HeLa cells were infected with either WT L2 or cdu1::GII strains and at various time points in the Ct infectious cycle crude cell lysates were analyzed by western blot to assess the relative abundance of Inc proteins. At 36 and 48 hpi, the levels of InaC protein were found to be undetectable in cells infected with cdu1::GII in comparison to cells infected with the WT L2 strain (Figures 3A and 3B). Conversely, IpaM protein levels exhibited a decrease at 36 and 48 hpi in cells infected with cdu1::GII (Figure 3C and 3D). We were not able to detect CTL0480 levels by western blot but were successful in following CTL0480 expression by indirect immunofluorescence (Figure 3E). The relative abundance of CTL0480 at inclusion membranes was not affected in cdu1::GII inclusions at 24 hpi. However, at 36 hpi, a subpopulation of cells lost CTL0480 immunoreactivity and by 48 hpi, inclusion membranes of the cdu1::GII strain were devoid of CTL0480 while CTL0480 was prominently detected at the inclusion membranes of WT L2 (Figures 3E and Supplemental Figure 5). As controls for the specificity of antibodies used for western blots and for indirect immunofluorescence, we included cells infected with Ct lacking ipaM (ipaM::GII, Meier et al., 2022), with an inaC nonsense mutant (M407, Kokes et al., 2015), and with Ct lacking ctl0480 (ctl0480::GII, Shaw et al., 2018). Overall, our results indicate that steady state protein levels of InaC and IpaM and CTL0480 localization at the inclusion membrane, especially at late stages of infection, are dependent on Cdu1, and that Cdu1 acts at different stages in the infection cycle.

Cdu1 stabilizes InaC, IpaM, and CTL0480.

(A) Western blot analysis of endogenous InaC in HeLa cells infected for 24, 36, and 48 hours with Wt Ct (L2 pBOMB), cdu1 null (cdu1::GII pBOMB), and inaC null (M407) strains. Ct Slc1 and human alpha tubulin were used to determine Ct burdens and equal loading of protein extracts respectively. Western blot images are representative of 2 independent experiments. (B) Quantification of InaC abundance (InaC western blot signal from (A)) normalized to Slc1 western blot signal (from panel A) in Hela cells infectedwith a cdu1 null strain, relative to normalized InaC abundance in Hela cells infected with Wt Ct. (C) Western blot analysis of endogenous IpaM in HeLa cells infected for 24, 36, and 48 hours with Wt Ct, cdu1 null, and ipam null (ipam::GII) strains. Western blot images are representative of 2 independent experiments. (D) Quantification of normalized IpaM abundance (from (C) in Hela cells infectedwith a cdu1 null strain, relative to normalized IpaM abundance in Hela cells infected with Wt Ct. (E) Localization of CTL0480 during Ct infection of HeLa cells at 24, 36, and 48 hpi. CTL0480 signal (green) co-localizes with the inclusion membrane protein IncA (magenta) at the Ct inclusion membrane. Arrowheads highlight cdu1 null inclusions lacking CTL0480 at 36 hpi. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. Quantification of MYPT1 localization at inclusion membranes can be found in Supplemental Figure 5

The acetylase activity of Cdu1 is required for Cdu1 to protect itself, InaC, and IpaM from polyubiquitination and proteasomal degradation

The crystal structures of Cdu1 bound to Ub or Coenzyme A indicated that the adenosine and phosphate groups of Coenzyme A make contact with a helix in variable region 3 (VR-3) of Cdu1 while the Ile36-patch of Ub binds to the opposite face of the same helix (Pruneda et al., 2018). Although Cdu1 catalyzes both of its DUB and acetylase (Act) activities with the same active site (Pruneda et al., 2018) the two activities of Cdu1 can be uncoupled by the amino acid substitution K268E in VR-3 which disrupts Coenzyme A binding required for Act activity and by the amino acid substitution I225A in the Ub-binding region of VR-3 required for DUB activity (Pruneda et al., 2018). These substitutions allowed us to test which of Cdu1’s enzymatic functions are required for the observed effects on protein stability. We generated Ct shuttle plasmids, expressing WT Cdu1, a catalytically inactive variant of Cdu1 lacking both DUB and Act activities (Cdu1C345A) (Pruneda et al., 2018), a Cdu1 DUB-deficient variant (Cdu1I225A), and a Cdu1 Act-deficient variant (Cdu1K268E). All Cdu1 constructs were expressed from the cdu1 endogenous promoter as 3X Flag epitope-tagged proteins.

Plasmids expressing each Cdu1 variant were transformed into the cdu1::GII mutant and the resulting strains used to infect HeLa cells for 36 and 48 hours. The levels of endogenous InaC and IpaM in cell extracts of infected cells were monitored by western blot analysis. At 36 hpi, InaC protein levels drastically decreased in cells infected with cdu1 null strains transformed with empty vector or expressing the catalytic inactive variant of Cdu1 (Cdu1C345A-Flag) (Figure 4A). Likewise, IpaM protein levels diminished at 48 hpi during infection with the same strains (Figure 4B). Both InaC and IpaM protein levels were restored to wild type levels in cdu1 null strains complemented with wild type Cdu1-Flag (Figures 4A and 4B). Unexpectedly, cells infected with a cdu1 null strain ectopically expressing the DUB deficient Cdu1 variant (Cdu1I225A-Flag) displayed wild type levels of InaC and IpaM while the Act deficient variant (Cdu1K268E-Flag) did not (Figures 4A and 4B). These results suggest that the acetylase activity of Cdu1 rather than its DUB activity is required for Cdu1’s ability to stabilize InaC and IpaM proteins.

The acetylase activity of Cdu1 is required to stabilize Cdu1, InaC, and IpaM.

(A) Western blot analysis of endogenous InaC and Cdu1-Flag catalytic variants expressed from a plasmid (pBOMB). HeLa cells were infected for 36 hours with WT Ct (L2 434 Bu pBOMB), a cdu1 null strain (cdu1::GII pBOMB), and cdu1 null strains expressing wild type Cdu1-Flag and the Cdu1 variants C345A-Flag (catalytic inactive), I225A-Flag (DUB deficient), and K268E-Flag (Act deficient). Cdu1-Flag variants were expressed from a pBOMB shuttle plasmid. Protein lysates from HeLa cells infected with an inaC null (M407) strain were used to control for the specificity of anti-InaC antibodies. Western blot images are representative of two independent experiments. (B) Western blot analysis of endogenous IpaM and Cdu1-FLAG variants in crude extracts of HeLa cells infected for 48 hours with the same strains as describe in (A). Infection of HeLa cells with ipaM::GII was used to test for the specificity of the anti-IpaM antibody. Western blot images are representative of two independent experiments. (C) Western blot analysis of Cdu1C345A-Flag (catalytic inactive) and Cdu1K268E-Flag (Act deficient) expressed in a cdu1 null strain or WT Ct (L2 434 Bu) background after infection of HeLa cells for 24 hours. Both Cdu1 variants are stabilized by Cdu1 expressed in WT Ct. (D) Western blot analysis of Cdu1-Flag and Cdu1C345A-Flag expressed in a cdu1 null strain and Cdu1C345A-Flag expressed in WT Ct (L2 434 Bu) following immunoprecipitation (anti-Flag) from HeLa cell extracts after infection for 24 hours and treatment with MG132 (25 μM, 5 hours). Western blot image is a representative blot from at least three independent experiments. (E) Western blot analysis of endogenous InaC and IpaM following immunoprecipitation of inclusion membrane enriched subcellular fractions (24hpi) with anti acetylated lysine antibodies. Western blot image is representative of two independent experiments. (F) Western blot analysis (Flag WB) of acetylated lysine immunoprecipitates generated from inclusion membrane enriched subcellular fractions (40 hpi) derived from HeLa cells infected with WT Ct strains expressing Cdu1-Flag, CTL0480-Flag, or IpaM-Flag. Western blot image is representative of two independent experiments. (G) WB of acetylated lysine immunoprecipitates of inclusion membrane enriched fractions (24 hpi) of HeLa cells infected with WT Ct and with an inaC null strain (M407) expressing InaC-Flag. Western blot image is representative of two independent experiments. (H) The initiator methionine, Lys10, Lys126, and Lys263 of Cdu1 are acetylated by 24 hpi. One tyrosine (Y) residue and multiple serine (S) and threonine (T) residues in Cdu1 are also phosphorylated during Ct infection of HeLa cells (24 hpi). Three PX(S/T)P MAPK phosphorylation consensus sequence motifs were identified in the proline rich domain of Cdu1. Modified residues were identified by quantitative LC MS/MS analysis of immunoprecipitated Cdu1-Flag across 3 independent biological replicates. TMD: Transmembrane domain. PR: Proline rich. PSTP: PX(S/T)P motifs. Ac: Acetylation. P: Phosphorylation. Numbering corresponds to amino acids in Cdu1 protein sequence.

When we monitored the stability of each Flag-tagged Cdu1 variant, we observed that the catalytically inactive variant of Cdu1 (C345A-Flag) was destabilized (Figures 4A and 4B-Flag WB). We reasoned that Cdu1 also protects itself from being targeted for degradation in infected cells. As with InaC and IpaM, the acetylase but not the DUB activity of Cdu1 was required for Cdu1’s stability (Figures 4A and 4B). Although the C345A (DUB-, Act-) and K268E (Act-) amino acid substitutions in Cdu1 do not destabilize Cdu1 expressed in E. coli (Pruneda et al., 2018), it was possible that these substitutions impacted the expression and/or folding of Cdu1 in Chlamydia. To determine if these Cdu1 mutants were inherently unstable, we expressed Cdu1C345A-Flag and Cdu1K268E-Flag in WT L2 in the presence of endogenous Cdu1. We found that each Flag-tagged variant was stabilized (Figure 4C), indicating that endogenous Cdu1 protected the catalytically-deficient Cdu1 variants in trans. In addition, western blot analysis of immunoprecipitated Cdu1-Flag and Cdu1C345A-Flag expressed in a cdu1::GII mutant showed that while WT Cdu1-Flag was not modified by Lys48-linked poly-ubiquitination, Cdu1C345A-Flag was robustly modified by Lys48-linked polyUb in the presence of the proteasome inhibitor MG132 (Figure 4D). Moreover, endogenous Cdu1 protected Cdu1C345A-Flag from Lys48-linked polyUb when Cdu1C345A-Flag was expressed in a wild type L2 background (Figure 4D). These results indicate that the loss of Cdu1 activity likely leads to its Lys48-linked poly-ubiquitination and subsequent proteasome-dependent degradation.

Because Cdu1 autoacetylates itself and its Act activity is directed towards lysines (Pruneda et al., 2018) we postulated that Cdu1 may stabilize proteins from degradation by acetylating lysine residues that are potential targets of ubiquitination. We tested this hypothesis by assessing whether Cdu1, InaC, CTL0480, and IpaM are acetylated at lysines during infection. Fractions enriched for inclusion membranes were isolated by sub-cellular fractionation from HeLa cells infected with wild-type L2 (24 hpi), and proteins acetylated at lysines were immunoprecipitated. Western blot analysis of acetyl-lysine immunoprecipitates indicated that InaC and IpaM, but not the Inc protein IncA, were acetylated (Figure 4E). Western blot analysis of anti-acetyl-lysine immunoprecipitates of inclusion membrane-enriched membrane fractions of HeLa cells infected with L2 expressing Cdu1-Flag (24 hpi), CTL0480-Flag (40 hpi), and IpaM-Flag (40 hpi) also showed that all three Flag tagged effectors were acetylated at lysines (Figure 4F). We also determined that Flag-tagged InaC expressed in an inaC null (M407) background was acetylated at lysines as determined by western blot analysis of anti-acetyl-lysine immunoprecipitates (Figure 4G). In addition, we identified acetylated forms of Cdu1 from mass spectrometric analysis of Flag immunoprecipitates derived from extracts of HeLa cells infected with L2 expressing Cdu1-Flag, (24 hpi) (Figure 4H).

Cdu1’s acetylase activity shields inclusions from ubiquitination but is not sufficient to protect against IFNγ mediated antimicrobial activity

We reasoned that the lysine acetylase activity of Cdu1 is a prominent mechanism by which Cdu1 protects client proteins (at 24 hpi), since loss of Cdu1 or expression of the acetylase deficient variant of Cdu1 (Cdu1K268E-Flag) leads to a marked increase in Ub immunostaining at or near the periphery of cdu1::GII inclusions (> 80% of inclusions) compared to HeLa cells infected with cdu1::GII strains complemented with wild type or DUB deficient (I225A-Flag) Cdu1 strains (Figures 5A, 5B, and Supplemental Figure 6A). Given that Cdu1 appears to localize exclusively at inclusion membranes we predicted that its activity would be spatially restricted to the inclusion periphery. We tested this premise by co-infecting HeLa cells with an incA null strain (M923 (IncAR197*), Kokes et al., 2015) and the cdu1::GII strain. IncA mediates homotypic fusion of inclusion membranes and loss of IncA results in the accumulation of multiple unfused inclusions in cells infected at high MOIs (Hackstadt et al., 1999; Suchland et al., 2000; Pannekoek et al., 2005) (Figure 5C). As expected, incA mutants which retain Cdu1 activity did not accumulate Ub at or near the periphery of inclusion membranes. In HeLa cells coinfected with both cdu1::GII and M923 (IncAR197*), cdu1::GII Cdu1C345A and M923, or cdu1::GII Cdu1K268E and M923, only the incA null inclusions were protected from ubiquitination (Figures 5C, 5D, and Supplemental Figure 6B). Based on these observations we conclude that the acetylase activity of Cdu1 protects proteins in cis and that this activity is constrained to the membrane of the pathogenic vacuole consistent with previous reports (Auer et al., 2020).

The DUB activity of Cdu1 is not required for blocking the ubiquitination of inclusion membranes and Cdu1 is not required for protection against IFNγ mediated cellular immunity.

(A) Representative images of Ct inclusions decorated with ubiquitin during infection of HeLa cells with a cdu1::GII strain for 24 hours. Representative images of infected Hela cells used for quantification of ubiquitin decorated inclusions in (B) are shown in Supplemental Figure 6A. Antisera against the membrane protein Cap1 (green) was used to mark inclusion membranes. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. (B) Quantification of ubiquitinated inclusions as shown in (A). The Ub fluorescent signal was used to determine the number of infected cells with Ub decorated inclusions. The total number of ubiquitinated inclusions was divided by the total number of inclusions analyzed (defined by Cap1 staining). 87%, 86%, and 91% of inclusions were decorated with Ub in HeLa cells infected with a cdu1 null strain and cdu1 null strains expressing Cdu1C345A (DUB-, Act-), and Cdu1K268E (Act-) variants respectively. Representative images (panel (A) and Supplemental Figure 6A) and quantification of ubiquitinated inclusions were obtained from inclusions imaged in 10 fields across 3 independent biological replicates for each strain. p values were determined by a student paired t-test. (C) Representative images of HeLa cells co-infected with cdu1::GII (cdu1Δ) and incA null (incAΔ, M923) strains at 24 hpi. IncA-deficient inclusions do not fuse with other inclusions. In co-infected cells, Cdu1 present on the inclusion membranes of incAΔ strains did not block ubiquitination events at or near the inclusion membranes of neighboring cdu1Δ strains (mCherry signal from pBOMB4-MCI plasmid). Representative images of Hela cells infected with strains quantified in (D) are shown in Supplemental Figure 6B. DNA stained with Hoechst is shown in blue. Scale bar: 10μm. (D) Quantification of cdu1Δ inclusions as shown in (C) in which ubiquitination events are observed in regions of cdu1Δ inclusion membranes that are in direct apposition to incAΔ inclusion membranes (junctions). The total number of cdu1Δ inclusions ubiquitinated at inclusion junctions was divided by the total number of inclusions analyzed (Cap1 staining). 66% and 61% of cdu1Δ inclusions were decorated with Ub at junctions in HeLa cells co-infected with incAΔ and cdu1Δ strains or with incAΔ and a cdu1Δ strain ectopically expressing Cdu1K268E (Act-) respectively. Representative images (panel (C) and Supplemental Figure 6B) and quantification of cdu1Δ inclusions ubiquitinated at junctions are derived from inclusions imaged in 6 fields across 3 independent biological replicates for each condition. p values were determined by a student paired t-test. (E) Quantification of Ct inclusion production during infection of unprimed and IFNγ-primed (100 U/mL) A549 cells at 24 hours post infection. Inclusions were quantified by high-content imaging analysis. Plot reflects inclusion counts across 9 fields of view and 3 independent biological replicates. Inclusion counts by each strain in unprimed A549 cells were set to 100%. Inclusion counts resulting from cdu1::GII and garD::GII strains were normalized to corresponding parental Ct inclusion (100%) counts in unprimed cells. P-values were calculated by 2 way ANOVA analysis. (F) Representative images of RNF213 localizing to inclusions of WT Ct, garD::GII, and cdu1::GII strains during infection of A549 cells primed with IFNγ (100 U/mL). (G) Quantification of RNF213 localizing to Ct inclusions during infection of unprimed and IFNγ-primed (100 U/mL) A549 cells at 24 hours post infection. Plot reflects inclusion counts across 6 fields and 3 independent biological replicates.

Recently, the Ct inclusion membrane protein GarD was identified as a Ct effector that shields Ct from γ-interferon mediated ubiquitination by the IFNγ-inducible human ubiquitin E3 ligase RNF213 (Walsh et al., 2022). Because Cdu1 also protects the Ct inclusion from ubiquitination, we tested if Cdu1 also plays a role in protecting Ct from IFNγ-induced cell immunity. A549 cells were pretreated with IFNγ (100 U/mL) and infected with WT Ct, cdu1::GII, or garD::GII strains. Infections with a garD::GII strain led to an approximate 90% decrease in the number of inclusions formed relative to infections with its parental WT Ct strain (Figure 5E) while infections with a cdu1::GII or its parental WT Ct strain showed a modest reduction in inclusion formation (approximately 26% and 12% respectively) (Figure 5E). These results suggest that Cdu1 likely does not play a role in protecting Ct from IFNγ-mediated cellular immunity.

We also tested whether RNF213 localizes to inclusions that lack Cdu1, as observed in IFNγ-primed A549 cells infected with garD::GII strains (Walsh et al., 2022). RNF213 did not localize to Ct inclusions when cells were infected with either WT Ct or cdu1::GII strains, regardless of whether or not the A549 cells were treated with IFNγ (Figures 5F and 5G). In contrast, RNF213 localized to approximately 37% of inclusions in cells infected with garD::GII mutants of unprimed A549 cells, and 81% in IFNγ treated cells (Figures 5F and 5G). Based on these results, we conclude that Cdu1 does not play a role in protecting Ct from IFNγ-induced antimicrobial activity.

Cdu1 is required for F-actin assembly and Golgi ministack repositioning around the Ct inclusion, and for MYPT1 recruitment to Ct inclusions

InaC is required for Ct to assemble F-actin scaffolds and to reposition Golgi mini stacks around the periphery of the inclusion membrane (Kokes et al., 2015; Wesolowski et al., 2017; Haines et al., 2021). Because Cdu1 regulates InaC levels, we predicted that cdu1 mutants would phenocopy inaC mutants. We quantified the number of inclusions surrounded by F-actin cages at 40 hpi. In cells infected with WT L2 (parental strain of M407 (inaC null), Nguyen and Valdivia., 2012; Kokes et al., 2015), approximately 25% of inclusions were surrounded by F-actin, consistent with previous observations (Chin et al., 2012; Kokes et al., 2015) (Figures 6A and 6B, and Supplemental Figure 7). The number of inclusions surrounded by F-actin decreased to approximately 7% in cells infected with an inaC null strain (M407) and increased to approximately 49% in HeLa cells infected with an inaC null strain (M407) complemented with wild type InaC (Figures 6A and 6B, and Supplemental Figure 7). Cells infected with cdu1::GII mutants transformed with an empty plasmid or expressing Cdu1C345A-Flag (DUB-Act-) and Cdu1K268E-Flag (Act-) resulted in approximately 8%,13%, and 10% of of F-actin positive inclusions respectively (Figures 6A and 6B, and Supplemental Figure 7). In contrast, cells infected with cdu1::GII mutants expressing Cdu1-Flag and Cdu1I225A-Flag (DUB-) led to a marked increase in F-actin inclusions (approximately 52% and 46% respectively) (Figures 6A and 6B, and Supplemental Figure 7). From these observations we conclude that the acetylase activity of Cdu1 is required for Ct to promote assembly of F-actin around the Ct inclusion likely through the stabilization of InaC.

Cdu1 is required for assembly of F-actin, Golgi ministack repositioning, and MYPT1 recruitment to the inclusion.

(A) Examples of representative images of F-actin (arrowheads) (green, Alexa FluorTM Phalloidin) assembled around the Ct inclusion (magenta, anti Cdu1 and Cap1 staining) in HeLa cells infected for 40 hours. Representative images for each strain analyzed can be found in Supplemental Figure 7. (B) Quantification of Ct inclusion surrounded by F-actin normalized to the total number of inclusions analyzed during infection of Hela cells at 40 hpi. Quantification of surrounding F-actin were obtained from inclusions imaged in 6 fields across 3 independent biological replicates. p values were determined by one-way ANOVAs with a Student-Newman-Keuls post hoc test. Strains used: WT L2 (Rif-R 434 Bu, parent of M407), M407 (inaC null strain) p2TK2, M407 p2TK2-InaC, WT L2 (434 Bu) pBOMB, cdu1::GII pBOMB, cdu1::GII pBOMB-Cdu1 Flag, cdu1::GII pBOMB-Cdu1C345A Flag, cdu1::GII pBOMB-Cdu1I225A Flag, and cdu1::GII pBOMB-Cdu1K268E Flag. (C) Sample representative images of Golgi (anti GM130 staining, magenta) around Ct inclusions in HeLa cells infected for 24 hours. Representative images for each strain analyzed can be found in Supplemental Figure 8. (D) Quantification of Golgi dispersal around the Ct inclusion during infection of Hela cells for 24 hpi. The length of Golgi dispersed around each Ct inclusion imaged was measured and normalized to the perimeter length of each inclusion (% Golgi dispersion around the inclusion). Golgi dispersal around Ct inclusions was quantified from inclusions imaged in 6 fields across 3 independent biological replicates. p values were determined by a student paired t-test. Strains analyzed were the same ones as mentioned in (B). (E) Representative images of MYPT1 (magenta) at Ct inclusions (green, anti-IncA staining). Arrowheads represent cdu1 null inclusions with low MYPT1 signal. DNA stained with Hoechst is shown in blue in panels C and E. Scale bar: 10μm. (F) Quantification of MYPT1 at Ct inclusions as shown in (E). Representative images in (E) and quantification of MYPT1 recruitment in (F) were obtained from inclusions imaged in 6 fields across 3 independent replicates. p values were determined by a student paired t-test.

We next quantified Golgi dispersal in infected HeLa cells at 24 hpi, a process that is also dependent on InaC (Kokes et al., 2015; Wesolowski et al., 2017). In Hela cells infected with an inaC null strain (M407) Golgi dispersal was limited to approximately 26% of the Ct inclusion perimeter. In contrast, cells infected with either its parental WT L2 or with an inaC null strain (M407) complemented with wild type InaC, the Golgi is dispersed around 45% of the inclusion perimeter (Figures 6C and 6D, and Supplmental Figure 8). Similarly, Golgi dispersal around inclusions during infection with WT L2 and in cdu1::GII mutants expressing wild type Cdu1-Flag or Cdu1I225A-Flag (DUB-) was approximately 43%, 41%, and 43% respectively (Figures 6C and 6D, and Supplmental Figure 8). In HeLa cells infected with cdu1::GII and cdu1::GII strains expressing Cdu1C345A-Flag (DUB-Act-), and Cdu1K268E-Flag (Act-), Golgi repositioning was restricted to approximately 24%, 23%, and 23% of inclusion perimeters respectively (Figures 6C and 6D, and Supplemental Figure 8). These results confirm that both InaC and Cdu1 are required for efficient repositioning of the Golgi around the Ct inclusion as previously reported (Kokes et al., 2015; Wesolowski et al., 2017; Pruneda et al., 2018; Auer et al., 2020) and that this process is independent of Cdu1’s DUB activity but requires its acetylase activity. Moreover, our results suggest that Cdu1 promotes Golgi repositioning by protecting InaC-mediated redistribution of the Golgi around the Ct inclusion.

CTL0480 promotes recruitment of the myosin phosphatase subunit MYPT1 to the inclusion membrane where it regulates the extrusion of intact inclusions from host cells (Lutter et al., 2013; Shaw et al., 2018). Consistent with the gradual loss of CTL0480 from inclusions in cells infected with the cdu1::GII strain starting at 36 hpi (Figures 3E and Supplemental Figure 5) we also observed a complete loss of MYPT1 recruitment to inclusions by 48 hpi (Figures 6E and 6F).

Cdu1, InaC, IpaM, and CTL0480 are required for optimal extrusion of Ct from host cells

In the absence of Cdu1, the levels of InaC, CTL0480, and IpaM decreased late in infection (36 hpi and 48 hpi, Figures 3 and 4) suggesting that a prominent role of Cdu1 is to protect these Incs from degradation late in infection. At the end of its developmental cycle, Chlamydia exits host cells by promoting cellular lysis or by extrusion of intact inclusions (Hybiske and Stephens., 2007). Ct host cell exit by extrusion is an active process requiring a remodeling of the actin cytoskeleton and the function of Inc proteins (Hybiske and Stephens., 2007; Chin et al., 2012; Lutter et al., 2013; Shaw et al., 2018; Nguyen et al., 2018). CTL0480 recruits MYPT1 (an inhibitor of Myosin II motor complexes) to the inclusion membrane which prevents premature extrusion of Ct inclusions and loss of CTL0480 leads to increased rates of extrusion by Ct from infected HeLa cells (Lutter et al., 2013; Shaw et al., 2018) (Figure 7). Actin polymerization is also required for Ct extrusion (Hybiske and Stephens., 2007; Chin et al., 2012) suggesting that InaC dependent recruitment of F-actin to the inclusion may also contribute to optimal Ct extrusion. IpaM localizes to microdomains in the inclusion membrane that are proposed to function as foci for extrusion (Nguyen et al., 2018). Based on these observations, we postulated that Cdu1-mediated protection of CTL0480, InaC, and IpaM regulates the extrusion of Ct inclusions. We quantified the number of extrusions released from infected HeLa cells at 52 hpi and observed a 60% reduction in the number of extrusions in HeLa cells infected with the cdu1::GII strain relative to cells infected with WT L2 (Figures 7A and 7B). Complementation of cdu1::GII with either wild type Cdu1-Flag or Cdu1I225A-Flag (DUB-) restored extrusion production to near wild type levels. In contrast, HeLa cells infected with cdu1::GII mutants expressing Cdu1K268E-Flag (Act-), or inaC (inaC::GII, Wesolowski et al., 2017) and ipaM (ipaM::GII, Meier et al., 2020) null strains led to a 42%, 75%, and 58% reduction in extrusion production respectively (Figures 7A and 7B). The decrease in the number of extruded inclusions by these strains was not attributed to defects in inclusion biogenesis as they produced a comparable number of inclusions at 48 hpi relative to cells infected with WT L2 (Supplemental Figure 9). Consequently, we infer that InaC, IpaM, and Cdu1 collectively contribute to the promotion of optimal extrusion of Ct inclusions from host cells, with Cdu1 playing a central regulatory role by protecting these effectors from degradation.

Cdu1, InaC, and IpaM are required for optimal extrusion of Ct inclusions from HeLa cells.

(A) Representative images of extrusions isolated from HeLa cell monolayers infected with Ct strains for 52 hours. Scale bar: 200 μm (B) Quantification of the number of extruded inclusions produced by infected HeLa cell monolayers. p values were determined by a student paired t-test. (C) Quantification of the size of extruded inclusions quantified in (B). Extruded inclusions varied in size among cells infected with wild type L2 (average :43 μm), ipaM mutants (average: 56 μm) and cdu1 mutants complemented wild type Cdu1 (average: 52 μm) and Cdu1I225A (DUB-) (average: 60 μm). p values were determined by one-way ANOVAs with a Student-Newman-Keuls post hoc test. Representative images in (A) and quantification of extrusion number in (B) and extrusion size in (C) are based on images obtained from 6 fields across 3 independent biological replicates.

In contrast, infection of HeLa cells with a ctl0480::GII mutant strain led to an increase in the number of extruded inclusions as previously observed (Shaw et al., 2018) (Figures 7A and 7B). Therefore, even though the Cdu1-mediated protection of InaC and IpaM is important for the extrusion of inclusions and cdu1 mutants phenocopy the loss of InaC and IpaM, the phenotypic similarities do not extend to the increased number of extruded inclusions observed in cells infected with the ctl0480::GII mutant strain (Figures 7A and 7B). We infer from these observations that functions for both InaC and IpaM in the extrusion of inclusions are epistatic to CTL0480. Extruded inclusions produced during infection of HeLa cells also varied in size with an average diameter of 40 μm (Figures 7A and 7C). Interestingly, the loss of IpaM and over expression of Cdu1-Flag and Cdu1I225A-Flag (DUB-) shifted the size distribution of extrusions toward larger extrusions (Figures 7A and 7C) suggesting that Ct regulates the size of extruded inclusions through Cdu1.

Discussion

Several Chlamydia Inc proteins regulate interactions between the pathogenic vacuole and the host cytoskeleton, organelles, and vesicular trafficking pathways. These Inc proteins also modulate host cell death programs and promote Chlamydia exit from host cells (reviewed in Bugalhão and Mota., 2019). Given the central roles that Incs play in promoting Chlamydia intracellular infection, it is not surprising that they are targeted for inactivation by host cellular defenses. In response, Chlamydia has evolved mechanisms to protect Incs. In this study we show that the acetylase activity of the effector Cdu1 protects itself and three Inc proteins; InaC, IpaM, and CTL0480, from ubiquitination and degradation (Figure 8). Interestingly, all three Inc proteins play prominent roles in regulating the extrusion of inclusions from host cells (Figure 7). Observations that the encapsulation of Chlamydia within an extruded inclusion enhances survival of Chlamydia within macrophages (Zuck et al., 2017) together with the broad conservation of extrusion as an exit strategy among Chlamydia (Zuck et al., 2016) suggests that this mechanism is important for Chlamydia pathogenesis. Notably, a cdu1 mutant strain (cdu1::Tn, Fischer et al., 2017) displays reduced bacterial loads in a murine model of upper genital tract infections (Fischer et al., 2017). While neither this strain nor our cdu1::GII mutant strain shows evident growth impairments during infection of Hela cells (data not shown), the observed reduction in bacterial load in the absence of Cdu1 in animal models of infection could potentially stem from defects in extrusion production or from perturbations in Cdu1-dependent regulation of extrusion size. Thus, targeting Inc proteins that regulate extrusion for Ub-mediated destruction may be advantageous for the host. For instance, targeting InaC for degradation would limit F-actin dependent extrusions (Hybiske and Stephens., 2013; Chin et al., 2012) and InaC-dependent microtubule scaffolds around the inclusion (Wesolowski et al., 2017; Haines et al., 2021). CTL0480 functions as an inhibitor of extrusions through its role in modulating the activity of myosin light chain 2 (MLC2) (Lutter et al., 2013; Shaw et al., 2018). IpaM localizes to specialized microdomains in the inclusion membrane which are also sites of enrichment for over 9 inclusion membrane proteins including Ctl0480 and MrcA, both of which are required for Chlamydia extrusion (Mital et al., 2010; Lutter et al., 2013; Nguyen et al., 2018). We also find that the loss of IpaM shifted the size distribution of extrusions towards larger inclusions (Figure 7). We speculate that heterogeneity in the size of extrusions might facilitate uptake of some extrusions by innate immune cells at infected mucosal sites to promote Chlamydia LGV dissemination to distal sites in the genital tract and avoid clearance of Chlamydia by other immune cells (Zuck et al., 2017).

A model for acetylation mediated protection of the Inc proteins InaC, IpaM, and Ctl0480 from degradation.

The cellular Ub machinery targets C. trachomatis effectors, including the Inc proteins InaC, IpaM, and CTL0480 for ubiquitination and subsequent protein degradation. C. trachomatis counters such defense mechanisms by translocating Cdu1 which protect itself and all three Inc proteins from being targeted for ubiquitination and degradation through its acetylase activity. Cdu1 protects InaC and enables the recruitment of F-actin scaffolds and Golgi ministacks to the inclusion perimeter and CTL0480 to recruit the Myosin II regulator MYPT1. All three inclusion proteins and Cdu1 promote extrusion and dissemination of C. trachomatis inclusions late in infection. Image generated with BioRender.com.

Effectors that modulate the activity of other translocated effectors are referred to as “metaeffectors”, a term coined by Kubori and colleagues after observing that the L. pneumophila effector LubX which functions as an E3 ligase, ubiquitinates the translocated effector SidH leading to its degradation (Kubori et al., 2010). Several other effector-metaeffector interactions have been described in L. pneumophila, Salmonella enterica, and Brucella abortus which regulate the activity of other effectors either directly or indirectly by modifying the same host target or cellular process (Kubori et al., 2010; Neunuebel et al., 2011; Jeong et al., 2015; Urbanus et al., 2016; Smith et al., 2020; Iyer and Das., 2021). In this context we propose that Cdu1 functions as a metaeffector in Ct to protect multiple effectors. We also observed that Cdu1 interactions with InaC, IpaM, and CTL0480 likely occur independently from each other and that the kinetics of degradation in the absence of Cdu1 varies for each Inc (Figure 3).

Our findings indicate that the DUB activity of Cdu1 was not required to protect InaC, IpaM, and CTL0480 from ubiquitination. Instead we find that Cdu1’s lysine acetylase activity is required to protect these Inc proteins and Cdu1 itself from ubiquitination. Indeed, we found that all three Incs and Cdu1 are acetylated at lysines in infected cells. However, we were unable to determine if lysine acetylation in all four proteins was dependent on Cdu1s’ Act activity or if these PTMs are protective. Why the DUB activity of Cdu1 is unable to compensate for loss of its Act activity remains unknown. It is possible that Cdu1, like other DUBs, is regulated by PTMs (Komander et al., 2009a). For instance, phosphorylation of human CYLD inhibits its DUB activity towards TRAF2 while phosphorylation of human USP8 inhibits its DUB activity toward EGFR (Reiley et al., 2005; Mizuno et al., 2007). Mass spectrometry analysis of immunoprecipitated Flag tagged Cdu1 expressed in Ct revealed that Cdu1 is phosphorylated at multiple serine and threonine residues (Figure 4H) as previously suggested (Zadora et al., 2019). We identified three PX(S/T)P MAPK phosphorylation consensus sequence motifs in the proline rich domain (PRD) of Cdu1, suggesting that MAPKs may regulate the DUB activity of Cdu1.

Cdu1 homologs are found in multiple Chlamydia species including C. trachomatis, C. muridarum, C. suis, C. psitacci, C. abortus, C. caviae, and C. felis but is notably absent in the genomes of C. pneumoniae and C. pecorum. The acquisition of a second deubiquitinase paralog (Cdu2) has also occurred in C. trachomatis, C. muridarum, and C. suis. In the genomes of all three species, cdu2 resides directly adjacent to cdu1; an arrangement that presumptively arose from a gene duplication event. Cdu2 is a dedicated ULP with deubiquitinating and deneddylating activities (Misaghi et al., 2006; Pruneda et al., 2016). New evidence suggest that both paralogues might not be functionally redundant. The crystal structure of Cdu2 has revealed differences in residues involved in substrate recognition between Cdu1 and Cdu2 and that each paralog might recognize polyUb chains differently (Hausman et al., 2020). The processivity rates for removal of terminal Ub from polyUb chains also differs between both isopeptidases with Cdu2 exhibiting limited trimming of polyUb as compared to Cdu1 (Hausman et al., 2020).

Moreover, Cdu2 lacks the proline rich domain found in Cdu1 which might be important for regulation of Cdu1 enzymatic activity. The presence of Cdu2 might also explain the low incidence of human and Ct proteins that were differentially ubiquitinated in the absence of Cdu1 (Figure 1). Whereas several Chlamydia species have acquired either one or two deubiquitinase paralogs, both C. pneumoniae and C. pecorum have not. Instead, both species have acquired an unrelated deubiquitinase (ChlaOTU) belonging to the OTU family of proteases (Makarova et al., 2000; Furtado et al., 2013). Curiously, ChlaOTU is also found in C. psitacci, C. abortus, C. caviae, and C. felis all of which encode only Cdu1 and is absent in C. trachomatis, C. muridarum, and C. suis, all of which encode Cdu1 and Cdu2. It is noteworthy that Chlamydia species have independently acquired deubiquitinases multiple times (Cdu1, Cdu2, ChlaOTU) and that some of these deubiquitinases have evolved into moonlighting enzymes reflecting the diverse strategies adopted by pathogenic Chlamydia as they adapt to their particular niche.

Supporting information